---
title: "18. Assumptions & Non-parametric tests"
author: "Sharon Kinkenberg & Johnny van Doorn"
date: "18 oct 2024"
footer: "Scientific & Statistical Reasoning"
format: 
  html: default
  revealjs: 
    output-ext: slide.html
---

## Assumptions
- Not an all-or-nothing case
- Assess the degree of the violation
- Especially vulnerable in low sample sizes
- Visualize (Q-Q plot, boxplots, descriptives)
- Sometimes there are cures, sometimes not...
- Don't trash your data, but investigate: why is this occuring?

## Homogeneity of variances
- Apply correction when possible (Welch for t-test/one-way ANOVA)
- Adjusts df $\rightarrow p$-value & CI
- Jane Superbrain 14.3 (pointless p)

## Sphericity
- Apply correction when possible (Greenhouse-Geisser)
- Adjusts df $\rightarrow p$-value & CI
- Jane Superbrain 14.3 (pointless p)

## Normality
- Shape of distribution
    - [Illustration of consequences](https://johnnydoorn.github.io/teaching-statistics/extra-texts/ViolatingNormalityANOVA.html)
- Outliers


```{r child="../../topics/non_parametric_concept/non_parametric_concept.qmd", eval=TRUE}
```

```{r child="../../topics/nonpar_wilcoxon_rank_sum_test/wilcoxon_rank_sum_test.qmd", eval=TRUE}
```

```{r child="../../topics/nonpar_wilcoxon_signed_rank_test/wilcoxon_signed_rank_test.qmd", eval=TRUE}
```

```{r child="../../topics/nonpar_kruskal_wallis_test/kruskal_wallis_test.qmd", eval=TRUE}
```

```{r child="../../topics/nonpar_friedman_anova/friedman_anova.qmd", eval=TRUE}
```

<!-- # ```{r child="../../../../topics/rank_correlation/rank_correlation.qmd", eval=TRUE} -->
<!-- # ``` -->

<!-- Footer insert below -->

```{r child="../../footer.qmd"}
```

