---
title: "22. Multiple regression"
author: "Johnny van Doorn"
date: "TODO 2026"
footer: "Scientific & Statistical Reasoning"
format:
  revealjs:
    output-ext: slide.html
    css: ../../ajohnson.css
  html: default
css: ../../ajohnson.css
---

# Block 3: Life is Continuous

- Regression with multiple predictors
- Interaction effects
- Mediation effects
- Chi-square
- Bayes!

---

<div class="goals">
In this lecture we aim to:

- Repeat some regression
- Look at multiple predictors
- Show these in JASP
    
<strong>Reading:</strong>
Chapter 8 (§8.9–8.16)
</div>


```{r child="regression_multiple_predictors.qmd"}
```


# Closing

## Recap

- In linear regression, we can build multiple models and look at their fit
- Just as in ANOVA, when we have multiple predictors we look out for:
  - Dependence between predictors
  - Interaction effects

## Recommended Exercises

- [Exercise 8.4](https://discoverjasp.com/pages/exercises#chapter-8)
- [Exercise 8.7](https://discoverjasp.com/pages/exercises#chapter-8)

```{r child="../../footer.qmd"}
```
