---
title: "6. Correlation"
author: "Johnny van Doorn"
date: "11 September 2026"
view: "https://amsuni-my.sharepoint.com/:v:/g/personal/j_b_vandoorn_uva_nl/IQDrIfwQyyl8TZPGFsiFTp7qAZjx_earKlylnJmV_gSG5eg?e=DYoMrE&nav=eyJyZWZlcnJhbEluZm8iOnsicmVmZXJyYWxBcHAiOiJTdHJlYW1XZWJBcHAiLCJyZWZlcnJhbFZpZXciOiJTaGFyZURpYWxvZy1MaW5rIiwicmVmZXJyYWxBcHBQbGF0Zm9ybSI6IldlYiIsInJlZmVycmFsTW9kZSI6InZpZXcifX0%3D"
footer: "Scientific & Statistical Reasoning"
format:
  revealjs:
    output-ext: slide.html
    css: ../../ajohnson.css
  html: default
css: ../../ajohnson.css
---

<div class="goals">
In this lecture we discuss:

* The t-distribution
  * Sampling distributions, continued
  * Standard error and confidence intervals
* Correlation
  * From variance to covariance to $r$
  * In JASP
* Significance of a correlation
  * Converting $r$ to a $t$-statistic
* Control for a third variable
  * Partial correlation
    
<strong>Reading:</strong>
Chapter 7 (§7.1–7.8), 5.7 (Scatterplots), 2.8–2.9 (Standard error &amp; confidence intervals)

<small>Not exam material: §7.2.5, §7.4.3–7.4.5 (Spearman, Kendall, point-biserial), §7.6 (comparing correlations).</small>
</div>


```{r child="challenger-recap.qmd"}
```


```{r child="t-distribution.qmd"}
```


```{r child="correlation.qmd"}
```


```{r child="correlation-partial.qmd"}
```


<!-- ```{r child="variance-standardization.qmd"} -->
<!-- ``` -->


# Closing

## Next Week

- The linear model
  - Predictions
  - Assumptions
  - Fun

## Recommended Exercises

- [Exercise 7.1](https://discoverjasp.com/pages/smart_alex#task-7.1)
  - Note: `pets.jasp` can be downloaded from the JASP data library
- [Exercise 7.6](https://discoverjasp.com/pages/smart_alex#task-7.6)
- [Exercise 7.7](https://discoverjasp.com/pages/smart_alex#task-7.7)

```{r child="../../footer.qmd"}
```
