---
title: "Correlation (next year draft)"
author: "Johnny van Doorn"
footer: "Scientific & Statistical Reasoning"
format:
  revealjs:
    output-ext: slide.html
    css: ../../ajohnson.css
  html: default
css: ../../ajohnson.css
---

<!-- Draft for NEXT year. The only change against the taught 2026 deck is that
     the Challenger example, which opens the lecture, now comes back at the end
     of the correlation section with its own r, t and p: the promise made in the
     recap ("one of them is significant and the other is not") gets paid off in
     the notation the lecture just built. Deliberately unnumbered so ssr.qmd
     does not list it. The taught version is 6. correlation.qmd, untouched.

     Only correlation.qmd is forked (as next-year-correlation.qmd); the other
     children are reused as-is so they cannot drift. Wrapper order matters:
     t-distribution.qmd defines tCurve(), which both the correlation child and
     the new Challenger slide call. -->

<div class="goals">
In this lecture we discuss:

* The t-distribution
  * Sampling distributions, continued
  * Standard error and confidence intervals
* Correlation
  * From variance to covariance to $r$
  * In JASP
* Significance of a correlation
  * Converting $r$ to a $t$-statistic
  * Back to Challenger
* Control for a third variable
  * Partial correlation

<strong>Reading:</strong>
Chapter 7 (§7.1–7.8), 5.7 (Scatterplots), 2.8–2.9 (Standard error &amp; confidence intervals)

<small>Not exam material: §7.2.5, §7.4.3–7.4.5 (Spearman, Kendall, point-biserial), §7.6 (comparing correlations).</small>
</div>


```{r child="challenger-recap.qmd"}
```


```{r child="t-distribution.qmd"}
```


```{r child="next-year-correlation.qmd"}
```


```{r child="correlation-partial.qmd"}
```


# Closing

## Next Week

- The linear model
  - Predictions
  - Assumptions
  - Fun

## Recommended Exercises

- [Exercise 7.1](https://discoverjasp.com/pages/smart_alex#task-7.1)
  - Note: `pets.jasp` can be downloaded from the JASP data library
- [Exercise 7.6](https://discoverjasp.com/pages/smart_alex#task-7.6)
- [Exercise 7.7](https://discoverjasp.com/pages/smart_alex#task-7.7)

```{r child="../../footer.qmd"}
```
