---
title: "5. JASP: Visualization and Correlation"
author: "Johnny van Doorn"
date: "TODO 2026"
footer: "Scientific & Statistical Reasoning"
format:
  revealjs:
    output-ext: slide.html
    css: ../../ajohnson.css
  html: default
css: ../../ajohnson.css
---

<div class="goals">
In block 1 we aim to:

* Discuss NHST concepts
* Introduce JASP
* Correlation
* Linear model (regression)
    
<strong>Reading:</strong>
Chapters 1–8.8
</div>

--- 

<div class="goals">
In this lecture we aim to:

* Introduce JASP
  * Descriptive statistics
  * Data visualizations
* Correlation
    
<strong>Reading:</strong>
Chapters 4 (§4.1–4.10), 5 (§5.1–5.6), 7 (§7.1–7.7)
</div>


```{r child="JASP.qmd", eval=TRUE}
```


```{r child="visualization.qmd", eval=TRUE}
```


```{r child="correlation.qmd", eval=TRUE}
```


# Closing

## Next Week

- Partial correlation
- The linear model
  - Predictions
  - Assumptions
  - Fun

## Recommended Exercises

- [Exercise 5.1](https://discoverjasp.com/pages/exercises#chapter-5)
- [Exercise 7.1](https://discoverjasp.com/pages/exercises#chapter-7)
  - Note: `pets.jasp` can be downloaded from the JASP data library

```{r child="../../footer.qmd"}
```
