---
title: "Lecture 6 — Reporting Guidelines & Exam Preparation"
author: "Johnny van Doorn"
date: "2026-06-17"
format:
  html:
    css: ajohnson.css
    code-fold: true
    theme: cosmo
    toc: true
    toc-depth: 3
    number-sections: false
    df-print: paged
    self-contained: true
  revealjs:
    output-file: lecture6_examprep_slides.html
    theme: simple
    css: ajohnson.css
    slide-number: true
    scrollable: true
    smaller: true
    df-print: kable
    embed-resources: true
engine: knitr
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(bayesrules)
library(ggplot2)
```

<div class="goals">
By the end of this lecture, you will be able to:

- Apply the van Doorn et al. (2021) reporting checklist to a Bayesian analysis
- Use calibrated language when describing BFs and posteriors
- Identify common errors in written Bayesian reports
- Demonstrate the full analysis pipeline on exam-style questions
</div>

<div class="note">
**Reading:** van Doorn et al. (2021), *Psychonomic Bulletin & Review*. doi: [10.3758/s13423-020-01798-5](https://doi.org/10.3758/s13423-020-01798-5)

**Exam:** Friday, June 19, 09:00–11:00, IWO 4.04C (Blauw). One A4 cheat sheet allowed.
</div>

---

# Reporting Guidelines (van Doorn et al., 2021)

## Why Reporting Matters

A Bayesian analysis is only as good as its communication. Two researchers with the same data but different priors can legitimately reach different conclusions — that's fine, but readers must be able to evaluate and reproduce the analysis.

Van Doorn et al. identify **four stages** of a Bayesian study:

$$\text{Planning} \;\rightarrow\; \text{Executing} \;\rightarrow\; \text{Interpreting} \;\rightarrow\; \text{Reporting}$$

## Stage 1: Planning

<div class="describe">

**What to decide before you see the data**

- Specify the **goal**: estimation, hypothesis testing, or both?
- If testing: one-sided or two-sided? Justify *theoretically* (before seeing data).
- Choose a **statistical model**: likelihood + prior family + prior parameters.
- Plan a **robustness check**: which alternative priors will you report?
- Specify a **sampling plan** (optional for Bayesians, but transparent).
- **Pre-register** for confirmatory research.

</div>

## Stage 2: Executing

- Check assumptions before running (outliers, distributional fit).
- Run the pre-registered analysis; **annotate any deviations**.
- Conduct the robustness check — vary the prior and report how the BF changes.

<div class="note">
A Bayesian analysis does not safeguard against model misspecification. Anscombe's quartet applies here too — always inspect your data.
</div>

## Stage 3: Interpreting

<div class="describe">

**For hypothesis testing (BF):**

- BF is a *relative* measure: data are $x$ times more likely under $H_1$ than $H_0$
- $BF_{10} = 1/BF_{01}$ — always use the subscript to be unambiguous
- BF does **not** equal posterior probability of $H_1$; that requires prior odds:
$$\frac{P(H_1 \mid y)}{P(H_0 \mid y)} = BF_{10} \times \frac{P(H_1)}{P(H_0)}$$

**For estimation (posterior):**

- Report posterior median (or mean) and a credible interval (ETI or HDI — specify which)
- The CI has a direct probability interpretation: $P(\theta \in [L,U] \mid y) = 0.95$

</div>

## Stage 4: Reporting — The Complete Checklist

**Model specification**

- [ ] State the likelihood explicitly
- [ ] State the prior with all parameters; justify the choice
- [ ] Report the prior ESS

**Checking**

- [ ] Prior predictive check: are the prior predictions plausible?
- [ ] Posterior predictive check: does the model fit the data?

**Results**

- [ ] Posterior median/mean and credible interval (ETI or HDI; state which)
- [ ] If testing: report $BF_{10}$ (or $BF_{01}$) with both hypotheses written out explicitly
- [ ] Sensitivity analysis: repeat under 2–3 alternative priors

**Transparency**

- [ ] Share data, code, and prior specification (OSF, GitHub, or supplementary)
- [ ] Use calibrated language (see below)

## Language Guide

<div style="display: flex; gap: 2em;">

<div style="flex: 1;">

**Use these**

- "Moderate evidence for $H_1$ ($BF_{10} = 7.2$)"
- "The data are $x$ times more likely under $H_1$"
- "The posterior mean is 0.65 (95% ETI: [0.51, 0.78])"
- "Evidence against the null ($BF_{01} = 12$)"
- "Anecdotal support for $H_1$"

</div>

<div style="flex: 1;">

**Avoid these**

- "The null hypothesis is false / rejected"
- "We have proven that the effect exists"
- "$\theta$ is definitely in [0.51, 0.78]"
- "No evidence, so the null is true"
- "The BF confirms our hypothesis"
- "The CI shows significance"

</div>

</div>

## A Specimen Paragraph

> We modelled the probability of correct beer identification ($\theta$) with prior $\text{Beta}(6, 6)$ (ESS = 12, symmetric, no directional expectation). After observing $y = 42$ correct identifications out of $n = 57$, the posterior was $\text{Beta}(48, 21)$, with median 0.696 (95% ETI: [0.578, 0.806]). A Bayes factor of $BF_{10} = 58.3$ indicated very strong evidence that performance exceeds chance ($\theta = 0.5$). A sensitivity analysis under $\text{Beta}(1,1)$ yielded $BF_{10} = 43.1$, supporting the robustness of this conclusion. Data and analysis code are available at [OSF link].

---

# Five Common Exam Mistakes

## Mistake 1: Misreading the Credible Interval

<div class="note">

**Wrong:** "There is a 95% probability that the true $\theta$ is between 0.51 and 0.78 *in repeated sampling*."

**Correct:** "**Given the data and the prior**, there is a 95% probability that $\theta$ is between 0.51 and 0.78."

The frequentist "repeated sampling" clause belongs to the confidence interval, not the Bayesian credible interval.

</div>

## Mistake 2: BF = Posterior Probability

<div class="note">

**Wrong:** "$BF_{10} = 9$ means there is a 90% chance that $H_1$ is true."

**Correct:** "$BF_{10} = 9$ means the data are 9 times more likely under $H_1$ than under $H_0$. Converting to posterior probability requires specifying the prior odds."

$$P(H_1 \mid y) = \frac{BF_{10} \times P(H_1)}{BF_{10} \times P(H_1) + P(H_0)}$$

</div>

## Mistake 3: No Evidence ≠ Evidence for the Null

<div class="note">

**Wrong:** "$BF_{10} = 1.2$ — there is no effect."

**Correct:** "$BF_{10} = 1.2$ is anecdotal. The data were **uninformative**: they did not substantially shift beliefs in either direction."

Contrast with $BF_{01} = 14$: *this* is positive evidence for $H_0$.

</div>

## Mistake 4: Using the CI to Test a Hypothesis

<div class="note">

**Wrong:** "The 95% CI excludes 0.5, so $H_0: \theta = 0.5$ is rejected."

**Correct:** Credible intervals are for estimation. Use the Bayes factor for hypothesis testing. The CI and the BF answer different questions.

| Question | Tool |
|---|---|
| Does an effect exist? | Bayes factor |
| How large is the effect? | Posterior + CI |
| What data should we expect next? | Posterior predictive |

</div>

## Mistake 5: "No Justification Needed" for a Flat Prior

<div class="note">

**Wrong:** "We used Beta(1,1) since it is objective and requires no justification."

**Correct:** Beta(1,1) has ESS = 2 — it asserts that a success rate of 0.001 is just as plausible as 0.999. For a well-studied phenomenon, this is a real and often implausible modelling assumption. All priors require justification.

</div>

---

# Error-Spotting Exercises

## Report 1: Mindfulness & Memory

> We tested a mindfulness intervention on working memory ($n = 15$). $BF_{10} = 4.3$, so **we conclude the null is false**. The CI [39.8, 42.6] **proves the intervention is effective**. No robustness analysis was needed since results are clear.

::: {.panel-tabset}

### Exercise

Find **3 errors** in this report.

### Solution

1. **"Null is false"**: $BF_{10} = 4.3$ is moderate evidence; it cannot falsify $H_0$. Report as "moderate evidence in favour of $H_1$."
2. **"CI proves effectiveness"**: credible intervals summarise uncertainty, they cannot "prove" anything. We would also need to compare the CI to a meaningful null value.
3. **"No robustness needed"**: robustness checks are especially important when making strong claims, not an optional extra.

:::

## Report 2: Facial Feedback

> One-sided Bayesian $t$-test ($n_1 = 53$, $n_2 = 57$). Informed prior: $t(0.35, 0.102, 3)$. Result: $BF_{0-} = 11.5$. **This means $H_0$ is 11.5× more probable than $H_-$.** Data and code stored privately.

::: {.panel-tabset}

### Exercise

Find **2 errors** in this report.

### Solution

1. **"$H_0$ is 11.5× more probable"**: $BF_{01}$ is a likelihood ratio, not a probability ratio. Converting to probability ratios requires specifying prior odds.
2. **"Stored privately"**: Bayesian results depend critically on prior specification. Without access to data and code, readers cannot evaluate or reproduce the analysis.

:::

---

# Mock Exam

<div class="note">
Close laptops. Work in pairs. 20 minutes for all 5 questions. Solutions follow each question.
</div>

## Mock Q1: Identify the Model

A memory researcher measures the proportion of words correctly recalled from a list of 30. She expects recall to be around 60%, based on prior studies.

**a.** Which Bayesian model is appropriate? Write the data model and prior.

**b.** A participant recalls 21 words. Write the posterior.

**c.** Compute the posterior mean and interpret it relative to the prior mean.

::: {.panel-tabset}

### Question

Attempt before looking at the solution.

### Solution

**a.** $Y \mid \theta \sim \text{Binomial}(30, \theta)$; prior $\theta \sim \text{Beta}(12, 8)$ (ESS = 20, mean = 0.60).

**b.** $\theta \mid y = 21 \;\sim\; \text{Beta}(12 + 21,\; 8 + 9) = \text{Beta}(33, 17)$.

**c.**

```{r q1}
alpha_post <- 33; beta_post <- 17
post_mean <- alpha_post / (alpha_post + beta_post)
cat("Posterior mean:", round(post_mean, 3), "\n")
cat("Prior mean: 0.60 | Sample proportion: 21/30 =", round(21/30, 3), "\n")
```

The posterior mean (0.66) is between the prior mean (0.60) and the observed proportion (0.70). With ESS = 20 and $n = 30$, the data have slightly more weight than the prior.

:::

## Mock Q2: Posterior Summaries

Using $\theta \mid y \sim \text{Beta}(33, 17)$:

**a.** R code for the 95% ETI.

**b.** $P(\theta > 0.70 \mid y)$.

**c.** Correct this statement: *"We're 95% confident the true recall rate falls in this interval in repeated experiments."*

::: {.panel-tabset}

### Question

Attempt before looking at the solution.

### Solution

**a.**
```{r q2a}
qbeta(c(0.025, 0.975), shape1 = 33, shape2 = 17)
```

**b.**
```{r q2b}
pbeta(0.70, shape1 = 33, shape2 = 17, lower.tail = FALSE)
```

**c.** "Given the data and the prior, there is a 95% probability that $\theta$ is in this interval." The repeated-experiments clause belongs to frequentist confidence intervals, not Bayesian credible intervals.

:::

## Mock Q3: Bayes Factor

Prior Beta(12, 8); posterior Beta(33, 17).

**a.** Compute $BF_{10}$ for $H_0: \theta = 0.5$ using the Savage–Dickey density ratio.

**b.** Interpret using the Jeffreys scale.

**c.** Compute the one-sided $BF_{+0}$ for $H_+: \theta > 0.5$.

::: {.panel-tabset}

### Question

Attempt before looking at the solution.

### Solution

**a.**
```{r q3a}
prior_dens <- dbeta(0.5, 12, 8)
post_dens  <- dbeta(0.5, 33, 17)
BF10 <- prior_dens / post_dens
cat("BF_10 =", round(BF10, 2), "\n")
```

**b.** Using the Jeffreys scale: BF 10–30 is "strong evidence," 3–10 is "moderate." Report accordingly.

**c.**
```{r q3c}
p_prior  <- pbeta(0.5, 12, 8,  lower.tail = FALSE)
p_post   <- pbeta(0.5, 33, 17, lower.tail = FALSE)
BF_plus0 <- (p_post / (1 - p_post)) / (p_prior / (1 - p_prior))
cat("BF(+,0) =", round(BF_plus0, 2), "\n")
```

The one-sided BF is larger because it concentrates prior probability on the side supported by the data.

:::

## Mock Q4: Posterior Prediction

Using posterior Beta(33, 17), predict recall for a new participant (30 words).

**a.** Simulation code.

**b.** $E(Y' \mid y)$ and $P(Y' \geq 25 \mid y)$.

**c.** Posterior predictive check on the original $y = 21$. Does the model fit?

::: {.panel-tabset}

### Question

Attempt before looking at the solution.

### Solution

**a & b.**
```{r q4ab}
set.seed(1)
N <- 10000
theta_post <- rbeta(N, shape1 = 33, shape2 = 17)
y_pred     <- rbinom(N, size = 30, prob = theta_post)

cat("E(Y' | y) =", round(mean(y_pred), 2), "\n")
cat("P(Y' >= 25 | y) =", round(mean(y_pred >= 25), 3), "\n")
```

**c.**
```{r q4c}
pp_pval <- mean(y_pred >= 21)
cat("PP p-value =", round(pp_pval, 3), "\n")
```

A pp p-value near 0.5 indicates good fit — $y_\text{obs} = 21$ falls near the centre of the predictive distribution.

:::

## Mock Q5: Spot the Errors

> We used Beta(1,1) — no justification needed since it is uninformative. We observed $y = 18$, $n = 25$, yielding posterior Beta(19, 8). The 95% credible interval [0.54, 0.89] **shows the null ($\theta = 0.5$) is rejected**. With $BF_{10} = 6.1$, **we are confident $H_1$ is true**. Strong evidence makes sensitivity analysis unnecessary.

::: {.panel-tabset}

### Question

Find **4 errors**.

### Solution

1. **"No justification needed"**: Beta(1,1) has ESS = 2 — a real modelling assumption with real consequences.
2. **"CI shows the null is rejected"**: use the BF for testing; the CI is for estimation.
3. **"Confident $H_1$ is true"**: $BF_{10} = 6.1$ is moderate evidence. It does not mean $H_1$ is true.
4. **"Sensitivity analysis not needed"**: sensitivity analysis is *especially* important for strong claims. One prior is never enough.

:::

---

# Exam Tips

## What to Expect

- **Duration:** 2 hours · **Format:** closed-book (one A4 cheat sheet)

**Typical question types:**

- Identify the appropriate model and write the prior + posterior
- Compute posterior summaries (mean, ETI) by hand or describe the R code
- Interpret a Bayes factor and posterior predictive
- Spot errors in a written Bayesian report
- Short conceptual questions (credible interval vs CI, BF vs posterior probability)

## Practical Tips

- **Show your reasoning**: partial credit is available. A correct setup with a computation error still earns marks.
- **Use the conjugate update rules**: memorise Beta→Beta, Gamma→Gamma, Normal→Normal.
- **Read the question carefully**: it will say whether to compute analytically, in R, or conceptually.
- **For BF questions**: state both hypotheses explicitly, then apply Savage–Dickey.
- **For reporting questions**: use the van Doorn et al. checklist as a scaffold.

## Suggested Cheat Sheet Contents

<div class="describe">

**Conjugate updates**

$\theta \mid y \sim \text{Beta}(\alpha+y,\;\beta+n-y)$

$\lambda \mid \mathbf{y} \sim \text{Gamma}(s+\sum y_i,\;r+n)$

$\mu_\text{post} = w_\text{pr}\,\mu_0 + w_\text{d}\,\bar{y}$

**Testing**

$\text{Post odds} = BF_{10} \times \text{Prior odds}$

$BF_{10} = f(\theta_0)\,/\,f(\theta_0 \mid y)$ (Savage–Dickey)

$BF_{+0} = \dfrac{P(\theta > \theta_0 \mid y)}{P(\theta > \theta_0)}$

**Credible interval:** direct probability statement — $P(\theta \in [L,U] \mid y) = 0.95$.

**ETI:** `qbeta(c(0.025, 0.975), a, b)` · **HDI:** `HPDinterval(as.mcmc(samples))`

</div>

---

<div class="note">
Open Q&A — any topic from the past three weeks. Exam is **Friday, June 19, 09:00–11:00, IWO 4.04C (Blauw)**. Good luck!
</div>
