---
title: "Lecture 6 — Reporting Guidelines & Exam Preparation"
author: "Johnny van Doorn"
date: "2026-06-17"
format:
  html:
    css: ajohnson.css
    code-fold: true
    theme: cosmo
    toc: true
    toc-depth: 3
    number-sections: false
    df-print: paged
    self-contained: true
  revealjs:
    output-file: lecture6_examprep_slides.html
    theme: simple
    slide-level: 3
    css: ajohnson.css
    slide-number: true
    scrollable: true
    smaller: true
    df-print: kable
    embed-resources: true
engine: knitr
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(bayesrules)
library(ggplot2)
```

<div class="goals">
By the end of this lecture, you will be able to:

- Apply the van Doorn et al. (2021) reporting checklist to a Bayesian analysis
- Use calibrated language when describing BFs and posteriors
- Identify common errors in written Bayesian reports
- Demonstrate the full analysis pipeline on exam-style questions
</div>

<div class="note">
**Reading:** van Doorn et al. (2021), *Psychonomic Bulletin & Review*. doi: [10.3758/s13423-020-01798-5](https://doi.org/10.3758/s13423-020-01798-5)

**Exam:** Friday, June 19, 09:00–11:00, IWO 4.04C (Blauw). One A4 cheat sheet allowed.
</div>

---

# Reporting Guidelines (van Doorn et al., 2021)

## Why Reporting Matters

A Bayesian analysis is only as good as its communication. Two researchers with the same data but different priors can legitimately reach different conclusions. That's fine, but readers must be able to evaluate and reproduce the analysis.

Van Doorn et al. identify **four stages** of a Bayesian study:

$$\text{Planning} \;\rightarrow\; \text{Executing} \;\rightarrow\; \text{Interpreting} \;\rightarrow\; \text{Reporting}$$

## Stage 1: Planning

<div class="describe">

**What to decide before you see the data**

- Specify the **goal**: estimation, hypothesis testing, or both?
- If testing: one-sided or two-sided? Justify *theoretically* (before seeing data).
- Choose a **statistical model**: likelihood + prior family + prior parameters.
- Plan a **robustness check**: which alternative priors will you report?
- Specify a **sampling plan** (optional for Bayesians, but transparent).
- **Pre-register** for confirmatory research.

</div>

## Stage 2: Executing

- Check assumptions before running (outliers, distributional fit).
- Run the pre-registered analysis; **annotate any deviations**.
- Conduct the robustness check: vary the prior and report how the BF changes.

<div class="note">
A Bayesian analysis does not safeguard against model misspecification. Anscombe's quartet applies here too; always inspect your data.
</div>

## Stage 3: Interpreting

<div class="describe">

**For hypothesis testing (BF):**

- BF is a *relative* measure: data are $x$ times more likely under $H_1$ than $H_0$
- $BF_{10} = 1/BF_{01}$; always use the subscript to be unambiguous
- BF does **not** equal posterior probability of $H_1$; that requires prior odds:
$$\frac{P(H_1 \mid y)}{P(H_0 \mid y)} = BF_{10} \times \frac{P(H_1)}{P(H_0)}$$

**For estimation (posterior):**

- Report posterior median (or mean) and a credible interval (ETI or HDI, specify which)
- The CI has a direct probability interpretation: $P(\theta \in [L,U] \mid y) = 0.95$

</div>

## Stage 4: Reporting (The Complete Checklist)

**Model specification**

- [ ] State the likelihood explicitly
- [ ] State the prior with all parameters; justify the choice
- [ ] Report the prior ESS

**Checking**

- [ ] Prior predictive check: are the prior predictions plausible?
- [ ] Posterior predictive check: does the model fit the data?

**Results**

- [ ] Posterior median/mean and credible interval (ETI or HDI; state which)
- [ ] If testing: report $BF_{10}$ (or $BF_{01}$) with both hypotheses written out explicitly
- [ ] Sensitivity analysis: repeat under 2–3 alternative priors

**Transparency**

- [ ] Share data, code, and prior specification (OSF, GitHub, or supplementary)
- [ ] Use calibrated language (see below)

## Language Guide

<div style="display: flex; gap: 2em;">

<div style="flex: 1;">

**Use these**

- "Moderate evidence for $H_1$ ($BF_{10} = 7.2$)"
- "The data are $x$ times more likely under $H_1$"
- "The posterior mean is 0.65 (95% ETI: [0.51, 0.78])"
- "Evidence against the null ($BF_{01} = 12$)"
- "Anecdotal support for $H_1$"

</div>

<div style="flex: 1;">

**Avoid these**

- "The null hypothesis is false / rejected"
- "We have proven that the effect exists"
- "$\theta$ is definitely in [0.51, 0.78]"
- "No evidence, so the null is true"
- "The BF confirms our hypothesis"
- "The CI shows significance"

</div>

</div>

## A Specimen Paragraph

> We modelled the probability of correct beer identification ($\theta$) with prior $\text{Beta}(6, 6)$ (ESS = 12, symmetric, no directional expectation). After observing $y = 42$ correct identifications out of $n = 57$, the posterior was $\text{Beta}(48, 21)$, with median 0.696 (95% ETI: [0.578, 0.806]). A Bayes factor of $BF_{10} = 58.3$ indicated very strong evidence that performance exceeds chance ($\theta = 0.5$). A sensitivity analysis under $\text{Beta}(1,1)$ yielded $BF_{10} = 43.1$, supporting the robustness of this conclusion. Data and analysis code are available at [OSF link].

---

# Five Common Exam Mistakes

## Mistake 1: Misreading the Credible Interval

<div class="note">

**Wrong:** "There is a 95% probability that the true $\theta$ is between 0.51 and 0.78 *in repeated sampling*."

**Correct:** "**Given the data and the prior**, there is a 95% probability that $\theta$ is between 0.51 and 0.78."

The frequentist "repeated sampling" clause belongs to the confidence interval, not the Bayesian credible interval.

</div>

## Mistake 2: BF = Posterior Probability

<div class="note">

**Wrong:** "$BF_{10} = 9$ means there is a 90% chance that $H_1$ is true."

**Correct:** "$BF_{10} = 9$ means the data are 9 times more likely under $H_1$ than under $H_0$. Converting to posterior probability requires specifying the prior odds."

$$P(H_1 \mid y) = \frac{BF_{10} \times P(H_1)}{BF_{10} \times P(H_1) + P(H_0)}$$

</div>

## Mistake 3: No Evidence ≠ Evidence for the Null

<div class="note">

**Wrong:** "$BF_{10} = 1.2$, there is no effect."

**Correct:** "$BF_{10} = 1.2$ is anecdotal. The data were **uninformative**: they did not substantially shift beliefs in either direction."

Contrast with $BF_{01} = 14$: *this* is positive evidence for $H_0$.

</div>

## Mistake 4: Using the CI to Test a Hypothesis

<div class="note">

**Wrong:** "The 95% CI excludes 0.5, so $H_0: \theta = 0.5$ is rejected."

**Correct:** Credible intervals are for estimation. Use the Bayes factor for hypothesis testing. The CI and the BF answer different questions.

| Question | Tool |
|---|---|
| Does an effect exist? | Bayes factor |
| How large is the effect? | Posterior + CI |
| What data should we expect next? | Posterior predictive |

</div>

## Mistake 5: "No Justification Needed" for a Flat Prior

<div class="note">

**Wrong:** "We used Beta(1,1) since it is objective and requires no justification."

**Correct:** Beta(1,1) has ESS = 2; it asserts that a success rate of 0.001 is just as plausible as 0.999. For a well-studied phenomenon, this is a real and often implausible modelling assumption. All priors require justification.

</div>

---

# Error-Spotting Exercises

## Report 1: Mindfulness & Memory

> We tested a mindfulness intervention on working memory ($n = 15$). $BF_{10} = 4.3$, so **we conclude the null is false**. The CI [39.8, 42.6] **proves the intervention is effective**. No robustness analysis was needed since results are clear.

::: {.panel-tabset}

### Exercise

Find **3 errors** in this report.

### Solution

1. **"Null is false"**: $BF_{10} = 4.3$ is moderate evidence; it cannot falsify $H_0$. Report as "moderate evidence in favour of $H_1$."
2. **"CI proves effectiveness"**: credible intervals summarise uncertainty, they cannot "prove" anything. We would also need to compare the CI to a meaningful null value.
3. **"No robustness needed"**: robustness checks are especially important when making strong claims, not an optional extra.

:::

## Report 2: Facial Feedback

> One-sided Bayesian $t$-test ($n_1 = 53$, $n_2 = 57$). Informed prior: $t(0.35, 0.102, 3)$. Result: $BF_{0-} = 11.5$. **This means $H_0$ is 11.5× more probable than $H_-$.** Data and code stored privately.

::: {.panel-tabset}

### Exercise

Find **2 errors** in this report.

### Solution

1. **"$H_0$ is 11.5× more probable"**: $BF_{01}$ is a likelihood ratio, not a probability ratio. Converting to probability ratios requires specifying prior odds.
2. **"Stored privately"**: Bayesian results depend critically on prior specification. Without access to data and code, readers cannot evaluate or reproduce the analysis.

:::

---

# Mock Exam

<div class="note">
Close laptops. Work in pairs. 20 minutes for all 3 questions. Solutions follow each question.
</div>

## Mock Q1: Conjugacy & Prior Choice (3.5 pts)
*(Lecture 3 / Ch. 5; cf. Beta–Binomial exercises)*

A food-delivery app wants to estimate $\theta$, the proportion of customers who tip their driver. Based on a pilot study, they adopt a $\text{Beta}(5, 5)$ prior. In the new cohort, **42 out of 60 customers** tip.

**a.** [1] Identify the data model and conjugate prior family. Write the posterior update rule and state the resulting posterior.

**b.** [1] Compute the prior mean, observed proportion, and posterior mean. Explain the influence of the prior vs data.

**c.** [0.75] Compute the 95% equal-tailed credible interval and interpret it.

**d.** [0.75] Compute $P(\theta > 0.70 \mid y)$.

::: {.panel-tabset}

### Question

Attempt before looking at the solution.

### Solution

**a.** $Y \mid \theta \sim \text{Binomial}(60, \theta)$; conjugate prior $\theta \sim \text{Beta}(\alpha,\beta)$. Update: $\theta \mid y \sim \text{Beta}(\alpha + y,\ \beta + n - y)$.

```{r m1a}
alpha <- 5; beta <- 5; y <- 42; n <- 60
ap <- alpha + y; bp <- beta + n - y
cat("Posterior: Beta(", ap, ",", bp, ")\n")
```

**b.**
```{r m1b}
cat("Prior mean:     ", round(alpha/(alpha+beta), 3), "\n")
cat("Observed prop:  ", round(y/n, 3), "\n")
cat("Posterior mean: ", round(ap/(ap+bp), 3), "\n")
cat("Prior ESS:      ", alpha + beta, "\n")
```

The posterior mean (0.70) is a weighted average of the prior mean (0.50) and observed proportion (0.70). With prior ESS = 10 vs n = 60, the data dominate.

**c.**
```{r m1c}
round(qbeta(c(0.025, 0.975), ap, bp), 3)
```

Given the data and prior, there is a 95% probability that $\theta$ lies in this interval.

**d.**
```{r m1d}
pbeta(0.70, ap, bp, lower.tail = FALSE)
```

:::

## Mock Q2: Grid Approximation (3 pts)
*(Lecture 4 / Ch. 6–7; cf. grid approximation exercises)*

A researcher models $\theta$, the probability of a new drug being effective. They use a $\text{Beta}(2, 2)$ prior **truncated to $[0.3, 0.7]$**. In a trial of **20 patients, 12 show improvement**.

**a.** [0.75] Explain why this truncated prior has no closed-form conjugate posterior, and outline the grid-approximation recipe.

**b.** [1] Implement the grid approximation (5,000 points on $[0,1]$) and report the posterior mean. Compare to the untruncated conjugate posterior mean.

**c.** [0.75] A random-walk Metropolis sampler on the **untruncated** model is at $\theta_c = 0.55$ and proposes $\theta_p = 0.60$. Using the unnormalised posterior, compute the Metropolis acceptance probability $\alpha$. Is the move accepted with certainty?

**d.** [0.5] What step size pathology causes: (i) a flat caterpillar trace plot; (ii) a drifting chain; (iii) a stuck chain?

::: {.panel-tabset}

### Question

Attempt before looking at the solution.

### Solution

**a.** The conjugate update requires a proper Beta prior on $[0,1]$. Truncation leaves the Beta family, so the posterior has no closed form. Grid approximation evaluates $u_i = f(\theta_i)L(y\mid\theta_i)$ on a fine grid, normalises to $u_i/\sum_j u_j$, and samples grid points proportional to those weights.

**b.**
```{r m2b}
y <- 12; n <- 20; grid <- seq(0, 1, length.out = 5000)
prior <- dbeta(grid, 2, 2) * (grid >= 0.3 & grid <= 0.7)
lik   <- dbinom(y, n, grid)
post  <- prior * lik; post <- post / sum(post)
cat("Grid posterior mean:        ", round(sum(grid * post), 3), "\n")
cat("Untruncated conjugate mean: ", round((2+y)/(4+n), 3), "\n")
```

Both ≈0.60. The truncation barely matters since data concentrate $\theta$ inside $[0.3, 0.7]$.

**c.**
```{r m2c}
logpost <- function(t) dbeta(t, 2, 2, log = TRUE) + dbinom(y, n, t, log = TRUE)
cat("Acceptance probability:", round(min(1, exp(logpost(0.60) - logpost(0.55))), 3), "\n")
```

$\alpha \approx$ [value]. If < 1, accept with that probability; if = 1, accept with certainty.

**d.** (i) Flat caterpillar = good mixing (optimal step size). (ii) Slow drift = step size **too small** (high autocorrelation). (iii) Stuck plateaus = step size **too large** (proposals overshoot and are rejected).

:::

## Mock Q3: Inference, Prediction & Testing (3.5 pts)
*(Lecture 5 / Ch. 8; cf. posterior-summary, prediction & Bayes-factor exercises)*

A clinic models $\lambda$, the mean number of emergency calls per hour, with a $\text{Gamma}(3, 1)$ prior. Over **8 hours** they record **15 calls**.

**a.** [0.75] Give the posterior distribution, its mean, and a 95% equal-tailed credible interval.

**b.** [1] Predict calls next hour. Compute $E(Y' \mid y)$ and $P(Y' \ge 2 \mid y)$ using the posterior-predictive (Negative-Binomial) distribution.

**c.** [1] Test $H_+: \lambda > 2$ vs $H_-: \lambda < 2$. Compute $BF_{+-}$ from prior and posterior odds and interpret.

**d.** [0.75] Test $H_0: \lambda = 2$ using Savage–Dickey. What does $BF_{01}$ say about $\lambda = 2$?

::: {.panel-tabset}

### Question

Attempt before looking at the solution.

### Solution

**a.** $Y_i\mid\lambda \overset{iid}{\sim}\text{Poisson}(\lambda)$; posterior $\lambda\mid y \sim \text{Gamma}(s+\sum y,\ r+n)$.

```{r m3a}
s <- 3; r <- 1; sumy <- 15; nobs <- 8
sp <- s + sumy; rp <- r + nobs
cat("Posterior: Gamma(", sp, ",", rp, "),  mean =", round(sp/rp, 3), "\n")
cat("95% ETI:", round(qgamma(c(0.025, 0.975), sp, rp), 3), "\n")
```

**b.**
```{r m3b}
cat("E(Y' | y) =",       round(sp/rp, 3), "\n")
cat("P(Y' >= 2 | y) =",  round(1 - pnbinom(1, size = sp, prob = rp/(rp+1)), 3), "\n")
```

The posterior-predictive marginalises over uncertainty in $\lambda$. Plugging in a point estimate ignores this uncertainty.

**c.**
```{r m3c}
prior_p <- pgamma(2, s, r); post_p <- pgamma(2, sp, rp)
BF_plusminus <- (post_p/(1-post_p)) / (prior_p/(1-prior_p))
cat("Prior P(l>2):", round(prior_p,3), " Posterior P(l>2):", round(post_p,3), "\n")
cat("BF(+, -) =", round(BF_plusminus, 2), "\n")
```

Interpret using Kass & Raftery scale.

**d.**
```{r m3d}
cat("BF01 =", round(dgamma(2, sp, rp) / dgamma(2, s, r), 3), "\n")
cat("BF10 =", round(dgamma(2, s, r) / dgamma(2, sp, rp), 3), "\n")
```

$BF_{01} > 1$ means posterior density at $\lambda=2$ exceeds prior density. Data support $\lambda=2$.

:::

---

# Exam Tips

## What to Expect

- **Duration:** 2 hours · **Format:** closed-book (one A4 cheat sheet)
- **Content:** Questions are re-skinned versions of *Bayes Rules!* end-of-chapter exercises, covering Lectures 3–5 (Ch. 5–8).

**Typical question types:**

- **Conjugate models** (Beta-Binomial, Gamma-Poisson, Normal-Normal): identify model, write prior and posterior, compute summaries
- **Grid approximation & MCMC**: implement grid approximation for non-conjugate priors, compute Metropolis acceptance probability, diagnose chain mixing
- **Posterior inference & prediction**: compute ETI/HDI, posterior predictive distribution, prediction intervals
- **Hypothesis testing**: Bayes factors (Savage-Dickey, prior/posterior odds), one-sided and two-sided tests, BF interpretation
- **Reporting & error-spotting**: identify common Bayesian reporting errors, apply van Doorn et al. checklist

## Practical Tips

- **Show your reasoning**: partial credit is available. A correct setup with a computation error still earns marks.
- **Use the conjugate update rules**: memorise Beta→Beta, Gamma→Gamma, Normal→Normal.
- **Read the question carefully**: it will say whether to compute analytically, in R, or conceptually.
- **For BF questions**: state both hypotheses explicitly, then apply Savage–Dickey.
- **For reporting questions**: use the van Doorn et al. checklist as a scaffold.

## Suggested Cheat Sheet Contents

<div class="describe">

**Conjugate updates**

$\theta \mid y \sim \text{Beta}(\alpha+y,\;\beta+n-y)$

$\lambda \mid \mathbf{y} \sim \text{Gamma}(s+\sum y_i,\;r+n)$

$\mu_\text{post} = w_\text{pr}\,\mu_0 + w_\text{d}\,\bar{y}$

**Testing**

$\text{Post odds} = BF_{10} \times \text{Prior odds}$

$BF_{10} = f(\theta_0)\,/\,f(\theta_0 \mid y)$ (Savage–Dickey)

$BF_{+0} = \dfrac{P(\theta > \theta_0 \mid y)}{P(\theta > \theta_0)}$

**Credible interval:** direct probability statement, $P(\theta \in [L,U] \mid y) = 0.95$.

**ETI:** `qbeta(c(0.025, 0.975), a, b)` · **HDI:** `HPDinterval(as.mcmc(samples))`

</div>

---

<div class="note">
Open Q&A: any topic from the past three weeks. Exam is **Friday, June 19, 09:00–11:00, IWO 4.04C (Blauw)**. Good luck!
</div>
