---
title: "8 Posterior Inference & Hypothesis Testing"
author: "Johnny van Doorn"
date: "2026-06-10"
format:
  html:
    css: ajohnson.css
    code-fold: true
    theme: cosmo
    toc: true
    toc-depth: 3
    number-sections: false
    df-print: paged
    self-contained: true
engine: knitr
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(bayesrules)
library(ggplot2)
library(gridExtra)
library(dplyr)
library(tidyr)
library(coda)
```

<div class="goals">
By the end of this lecture, you will be able to:

- Summarise a posterior distribution using point estimates and credible intervals (ETI and HPD)
- Evaluate interval hypotheses using posterior probabilities and Bayes factors
- Test point hypotheses using the Savage–Dickey density ratio
- Explain why and when to go beyond conjugacy (grid approximation, MCMC)
</div>

<div class="note">
**Recommended Reading:** *Bayes Rules!* Chapters 8.1–8.2; course addendum *Testing Point Hypotheses*. Suggested exercises: 2, 3, 4, 6, 7, 8, 9, 10, 14 (Ch. 8), and exercises 25–27 from the addendum.
</div>

---

# Review: Four Ingredients of a Bayesian Analysis

Every Bayesian model is built on the same four components, regardless of model complexity.

<div class="describe">

**The four ingredients**

1. **Prior model** — beliefs about parameter $\theta$ before seeing data: $f(\theta)$
2. **Data model (likelihood)** — how data arise given the parameter: $f(y \mid \theta)$
3. **Likelihood function** — the same expression viewed as a function of $\theta$ with $y$ fixed: $L(\theta \mid y) \propto f(y \mid \theta)$
4. **Posterior model** — updated beliefs after seeing data: $f(\theta \mid y) \propto f(\theta) \cdot L(\theta \mid y)$

</div>

The three conjugate families from last lecture:

| Likelihood | Prior | Posterior |
|---|---|---|
| Binomial$(n, \theta)$ | Beta$(\alpha, \beta)$ | Beta$(\alpha+y,\;\beta+n-y)$ |
| Poisson$(\lambda)$ | Gamma$(s, r)$ | Gamma$(s+\sum y_i,\; r+n)$ |
| Normal$(\mu, \sigma^2)$, $\sigma$ known | Normal$(\mu_0, \tau^2)$ | Normal$(\mu_\text{post}, \sigma^2_\text{post})$ |

**Today** we use these posteriors to draw conclusions: estimating parameters and testing hypotheses.

---

# Running Example: Dutch Railways

We will use one running example throughout this lecture.

<div class="describe">

**Scenario**

You are a data analyst for the Dutch Railway Service (NS). Your task: model the **monthly rate of signal failures** on a 10 km stretch of track.

- Parameter: $\lambda =$ mean failures per month
- Likelihood: $Y_i \mid \lambda \overset{\text{iid}}{\sim} \text{Poisson}(\lambda)$
- Prior: Gamma(2, 0.5) — failures are relatively rare, modest prior certainty (mean = 4)

</div>

```{r railways-setup}
library(bayesrules)

# Observed 4 months: y = (2, 0, 3, 1)
y <- c(2, 0, 3, 1)

# Posterior: Gamma(s + sum(y), r + n) = Gamma(8, 4.5)
s_post <- 2 + sum(y)
r_post <- 0.5 + length(y)
cat("Posterior: Gamma(", s_post, ",", r_post, ")\n")

plot_gamma_poisson(shape = 2, rate = 0.5, sum_y = sum(y), n = length(y))
summarize_gamma_poisson(shape = 2, rate = 0.5, sum_y = sum(y), n = length(y))
```

---

# Posterior Estimation (Ch. 8.1)

## Point Estimates

We can summarise the posterior with a single "best guess":

| Summary | Formula (Gamma) | Use when |
|---|---|---|
| **Mean** | $s / r$ | Symmetric posteriors; minimises squared error |
| **Mode (MAP)** | $(s-1) / r$ | Need the single most plausible value |
| **Median** | `qgamma(0.5, s, r)` | Skewed posteriors; more robust |

```{r point-estimates}
s <- 8; r <- 4.5

mean_lambda   <- s / r
mode_lambda   <- (s - 1) / r
median_lambda <- qgamma(0.5, shape = s, rate = r)
sd_lambda     <- sqrt(s) / r

cat("Mean:  ", round(mean_lambda, 3),   "\n")
cat("Mode:  ", round(mode_lambda, 3),   "\n")
cat("Median:", round(median_lambda, 3), "\n")
cat("SD:    ", round(sd_lambda, 3),     "\n")
```

<div class="note">
For symmetric posteriors (e.g. Normal–Normal), mean = median = mode. For skewed posteriors (e.g. Gamma), they differ — the mean is pulled toward the heavier tail. When in doubt about which to report, reporting all three is informative.
</div>

## Credible Intervals

<div class="describe">

**Posterior credible interval**

A $100(1-\alpha)\%$ credible interval $[L, U]$ satisfies:
$$P(L \leq \theta \leq U \mid y) = 1 - \alpha$$

This is a **direct probability statement** about the parameter given the data and prior. It says exactly what most people think a frequentist confidence interval says — but doesn't.

</div>

### Equal-Tailed Interval (ETI)

Cuts equal probability from each tail: $P(\theta < L \mid y) = P(\theta > U \mid y) = \alpha/2$.

```{r eti}
# 95% ETI for Gamma(8, 4.5)
ci_95 <- qgamma(c(0.025, 0.975), shape = s, rate = r)
ci_50 <- qgamma(c(0.25,  0.75),  shape = s, rate = r)
ci_99 <- qgamma(c(0.005, 0.995), shape = s, rate = r)

cat("95% ETI: [", round(ci_95[1], 3), ",", round(ci_95[2], 3), "]\n")
cat("50% ETI: [", round(ci_50[1], 3), ",", round(ci_50[2], 3), "]\n")
cat("99% ETI: [", round(ci_99[1], 3), ",", round(ci_99[2], 3), "]\n")
```

### Highest Density Interval (HDI)

The **shortest** interval containing $1-\alpha$ of the posterior mass — always includes the mode.

```{r hdi}
library(coda)
posterior_samples <- rgamma(1e5, shape = s, rate = r)
hdi_95 <- HPDinterval(as.mcmc(posterior_samples), prob = 0.95)
cat("95% HDI: [", round(hdi_95[1], 3), ",", round(hdi_95[2], 3), "]\n")
```

```{r eti-vs-hdi-plot, fig.width=10, fig.height=4}
lambda_vals <- seq(0, 4, length.out = 1000)
dens_df <- data.frame(x = lambda_vals,
                      y = dgamma(lambda_vals, shape = s, rate = r))

ci  <- qgamma(c(0.025, 0.975), shape = s, rate = r)
hdi_vals <- as.numeric(HPDinterval(as.mcmc(rgamma(1e5, s, r)), prob = 0.95))

make_panel <- function(lo, hi, title, fill_col) {
  ggplot(dens_df, aes(x, y)) +
    geom_line(linewidth = 1) +
    geom_area(data = subset(dens_df, x >= lo & x <= hi),
              aes(x, y), fill = fill_col, alpha = 0.4) +
    geom_vline(xintercept = (s-1)/r, linetype = "dashed",
               color = "tomato", linewidth = 1) +
    labs(title = title, x = expression(lambda), y = "Density") +
    theme_minimal(base_size = 13)
}

grid.arrange(
  make_panel(ci[1],       ci[2],       "95% ETI (equal-tailed)", "steelblue"),
  make_panel(hdi_vals[1], hdi_vals[2], "95% HDI (shortest)",     "darkorange"),
  ncol = 2
)
```

<div class="note">
**When to use which?**

- For **symmetric** posteriors (Normal–Normal): ETI ≈ HDI — either is fine.
- For **skewed** posteriors (Gamma, Beta near boundary): HDI is shorter and contains the mode. Prefer HDI when reporting the most credible parameter region.
- ETI always contains the median; HDI always contains the mode.
- Whichever you use, state which in your report.
</div>

---

# Hypothesis Testing (Ch. 8.2)

## Interval Hypotheses

An **interval hypothesis** asserts the parameter falls in some range. We evaluate it using the **Bayes factor**, which quantifies how much the data update our beliefs between two competing claims.

<div class="describe">

**Bayes factor for interval hypotheses**

$$
BF_{a0} = \frac{\text{Posterior odds}}{\text{Prior odds}} = \frac{P(H_a \mid y) / P(H_0 \mid y)}{P(H_a) / P(H_0)}
$$

- $BF_{a0} > 1$: data support $H_a$ over $H_0$
- $BF_{a0} < 1$: data support $H_0$ over $H_a$
- $BF_{a0} \approx 1$: data are uninformative about this contrast

</div>

### Dutch Railways: Is the failure rate below 2 per month?

$$H_a: \lambda < 2 \quad \text{vs.} \quad H_0: \lambda \geq 2$$

```{r interval-bf}
# Prior: Gamma(2, 0.5) — before seeing any data
prior_prob <- pgamma(2, shape = 2, rate = 0.5)
prior_odds <- prior_prob / (1 - prior_prob)
cat("Prior P(lambda < 2):", round(prior_prob, 3),
    "| Prior odds:", round(prior_odds, 3), "\n")

# Posterior after 4 months: Gamma(8, 4.5)
post_prob <- pgamma(2, shape = 8, rate = 4.5)
post_odds <- post_prob / (1 - post_prob)
cat("Posterior P(lambda < 2 | y):", round(post_prob, 3),
    "| Posterior odds:", round(post_odds, 3), "\n")

# Bayes factor
BF_a0 <- post_odds / prior_odds
cat("BF(a, 0) =", round(BF_a0, 3), "\n")
```

```{r interval-plot, fig.width=10, fig.height=5}
lambda_vals <- seq(0, 6, length.out = 500)
df <- data.frame(
  lambda    = lambda_vals,
  Prior     = dgamma(lambda_vals, shape = 2, rate = 0.5),
  Posterior = dgamma(lambda_vals, shape = 8, rate = 4.5)
)

thresh <- 2

make_hyp_panel <- function(data, col_name, fill_ha, title) {
  ggplot(data, aes(x = lambda, y = .data[[col_name]])) +
    geom_line(linewidth = 1) +
    geom_area(data = subset(data, lambda <  thresh),
              aes(y = .data[[col_name]]), fill = fill_ha, alpha = 0.4) +
    geom_area(data = subset(data, lambda >= thresh),
              aes(y = .data[[col_name]]), fill = "grey70", alpha = 0.3) +
    geom_vline(xintercept = thresh, linetype = "dashed", color = "grey30") +
    annotate("text", x = 0.6, y = max(data[[col_name]]) * 0.88,
             label = expression(H[a]), color = fill_ha, size = 5, fontface = "bold") +
    annotate("text", x = 3.5, y = max(data[[col_name]]) * 0.88,
             label = expression(H[0]), color = "grey40", size = 5, fontface = "bold") +
    labs(title = title, x = expression(lambda), y = "Density") +
    theme_minimal(base_size = 13)
}

grid.arrange(
  make_hyp_panel(df, "Prior",     "steelblue",  "Prior: Gamma(2, 0.5)"),
  make_hyp_panel(df, "Posterior", "darkorange", "Posterior: Gamma(8, 4.5)"),
  ncol = 2
)
```

**Interpretation:** A BF ≈ 5.8 indicates **moderate evidence** that the failure rate is below 2/month. The posterior shifted substantially toward $H_a$.

### Evidence scale (Kass & Raftery, 1995)

| BF | Evidence for $H_a$ |
|---|---|
| 1 – 3 | Barely worth mentioning |
| 3 – 10 | Moderate / substantial |
| 10 – 30 | Strong |
| 30 – 100 | Very strong |
| > 100 | Decisive |

### What happens with more data?

After 10 months ($\sum y = 15$), posterior becomes Gamma(17, 10.5):

```{r more-data-bf}
prior_odds_saved <- prior_prob / (1 - prior_prob)

post_prob2 <- pgamma(2, shape = 17, rate = 10.5)
post_odds2 <- post_prob2 / (1 - post_prob2)
BF_full    <- post_odds2 / prior_odds_saved

cat("BF after 10 months =", round(BF_full, 2), "\n")
```

The BF grows from 5.8 to ~14, crossing into **strong** evidence. More data, stronger conclusions.

---

## Point Hypotheses & the Savage–Dickey Density Ratio

A **point hypothesis** like $H_0: \lambda = 2$ cannot be evaluated using posterior probabilities — a continuous distribution assigns zero probability to any single point. Instead, we reframe as a **model comparison**:

- $H_0$: $\lambda$ is fixed at $\lambda_0 = 2$
- $H_1$: $\lambda$ is unknown, modelled with a continuous prior

<div class="describe">

**Savage–Dickey density ratio**

When $H_0$ is nested within $H_1$ (i.e. $H_0$ is a special case of the prior), the Bayes factor simplifies elegantly:

$$BF_{01} = \frac{f(\lambda_0 \mid y)}{f(\lambda_0)}$$

Compare the **posterior density** to the **prior density** at the null value. No integration needed.

- $BF_{01} > 1$: posterior peaked more at $\lambda_0$ than prior → evidence *for* $H_0$
- $BF_{01} < 1$: data moved probability away from $\lambda_0$ → evidence *against* $H_0$

</div>

```{r savage-dickey}
# Prior: Gamma(2, 0.5);  Posterior after 10 months: Gamma(17, 10.5)
lambda0 <- 2

prior_dens <- dgamma(lambda0, shape = 2,  rate = 0.5)
post_dens  <- dgamma(lambda0, shape = 17, rate = 10.5)

BF_01 <- post_dens / prior_dens
cat("Prior density at lambda=2:    ", round(prior_dens, 4), "\n")
cat("Posterior density at lambda=2:", round(post_dens,  4), "\n")
cat("BF_01 =", round(BF_01, 3), " (> 1: favours H_0)\n")
cat("BF_10 =", round(1/BF_01, 3), " (> 1: favours H_1)\n")
```

```{r savage-dickey-plot, fig.width=9, fig.height=4}
lambda_vals <- seq(0, 5, length.out = 500)
df_sd <- data.frame(
  lambda    = lambda_vals,
  Prior     = dgamma(lambda_vals, shape = 2,  rate = 0.5),
  Posterior = dgamma(lambda_vals, shape = 17, rate = 10.5)
)

df_long <- pivot_longer(df_sd, Prior:Posterior,
                        names_to = "Dist", values_to = "Density")

ggplot(df_long, aes(x = lambda, y = Density, color = Dist, fill = Dist)) +
  geom_area(alpha = 0.3, position = "identity") +
  geom_line(linewidth = 1.2) +
  geom_vline(xintercept = lambda0, linetype = "dashed", color = "grey30") +
  geom_point(aes(x = lambda0,
                 y = ifelse(Dist == "Prior",
                            dgamma(lambda0, 2, 0.5),
                            dgamma(lambda0, 17, 10.5))),
             size = 4, shape = 21, fill = "white") +
  annotate("text", x = lambda0 + 0.15, y = 0.55,
           label = paste0("BF[0*1] == ", round(BF_01, 2)),
           parse = TRUE, size = 5) +
  scale_color_manual(values = c("Prior" = "steelblue", "Posterior" = "darkorange")) +
  scale_fill_manual(values  = c("Prior" = "steelblue", "Posterior" = "darkorange")) +
  labs(title = "Savage–Dickey: comparing densities at the null value",
       x = expression(lambda), y = "Density") +
  theme_minimal(base_size = 13) +
  theme(legend.title = element_blank())
```

**Interpretation:** The posterior density at $\lambda = 2$ substantially exceeds the prior density — the data are *consistent with* the null benchmark. $BF_{01} \approx 3$ indicates modest evidence *for* $H_0: \lambda = 2$. This is nuanced: the interval test shows the rate is likely *below* 2, but the point test says it's not dramatically different from 2.

### Effect of prior diffuseness on the BF

Vaguer priors spread probability thinly, so they predict data less precisely everywhere — leading to larger $BF_{01}$ values for the null even when the data are not especially compatible with it. This is Occam's razor in action.

```{r sd-sensitivity}
# Same posterior (Gamma(17, 10.5)); compare tight vs diffuse prior
prior_tight   <- dgamma(lambda0, shape = 2, rate = 0.5)   # tight
prior_diffuse <- dgamma(lambda0, shape = 2, rate = 0.1)   # diffuse

cat("BF_01 with tight prior Gamma(2,0.5):  ", round(post_dens / prior_tight,   3), "\n")
cat("BF_01 with diffuse prior Gamma(2,0.1):", round(post_dens / prior_diffuse, 3), "\n")
```

<div class="note">
**Always** report a sensitivity analysis for Bayes factors. The BF depends on the prior under $H_1$; a diffuse alternative automatically yields larger $BF_{01}$ values.
</div>

---

# Beyond Conjugacy: Grid Approximation & MCMC

Conjugate models require prior and likelihood to match. Real research often involves:

- Multiple unknown parameters
- Non-standard prior shapes (truncated, mixture)
- Complex hierarchical models

The good news: $f(\theta \mid y) \propto f(\theta) \cdot L(y \mid \theta)$ always holds. We just need numerical methods to compute usable summaries.

## Grid Approximation

1. Define a grid: $\theta_1, \ldots, \theta_K$
2. Evaluate unnormalised posterior at each point: $u_i = f(\theta_i) \cdot L(y \mid \theta_i)$
3. Normalise: $\hat{f}(\theta_i \mid y) = u_i / \sum_j u_j$
4. Sample from the grid with probabilities proportional to $\hat{f}$

```{r grid-approx}
# Non-conjugate example: Beta(3,3) prior truncated to [0.4, 1]
theta_grid <- seq(0, 1, length.out = 2000)

prior      <- dbeta(theta_grid, 3, 3) * (theta_grid >= 0.4)
likelihood <- dbinom(18, size = 25, prob = theta_grid)

unnorm         <- prior * likelihood
posterior_grid <- unnorm / sum(unnorm)

theta_samples <- sample(theta_grid, size = 1e4,
                        replace = TRUE, prob = posterior_grid)

hist(theta_samples, breaks = 50, probability = TRUE,
     main = "Grid approx: truncated Beta(3,3) prior, y=18, n=25",
     xlab = expression(theta), col = "steelblue", border = "white")

cat("Posterior mean:", round(mean(theta_samples), 3), "\n")
cat("95% ETI: [", round(quantile(theta_samples, 0.025), 3),
    ",", round(quantile(theta_samples, 0.975), 3), "]\n")
```

**Limitation:** Grid points grow as $K^p$ with the number of parameters $p$. With 10 parameters and 100 grid points each, you need $100^{10} = 10^{20}$ evaluations — infeasible.

## MCMC — The Intuition

**Markov Chain Monte Carlo** draws samples from the posterior *without* computing the normalising constant.

<div class="describe">

**The island-hopping metaphor** (Metropolis algorithm)

A politician visits islands whose populations represent posterior density:

1. Propose a neighbouring island at random
2. If it's bigger → always move
3. If it's smaller → move with probability = (proposed / current population)
4. Otherwise → stay put

Long-run time spent on each island $\propto$ population — without ever needing the total.

</div>

In practice, use `brms` (which calls Stan). Key diagnostics to check:

| Term | Meaning |
|---|---|
| **Trace plot** | $\theta^{(t)}$ vs. $t$ — should look like a "hairy caterpillar" |
| **Burn-in** | Early samples discarded while chain finds the posterior |
| **$\hat{R}$** | Convergence — should be near 1.00 |
| **Effective N** | Consecutive samples are correlated; effective N < actual N |

Once you have MCMC draws, all posterior summaries work exactly the same as with conjugate samples:

```{r mcmc-use}
# theta_mcmc: any posterior sample (grid, MCMC, conjugate — identical workflow)
theta_mcmc <- theta_samples

mean(theta_mcmc)                        # posterior mean
median(theta_mcmc)                      # posterior median
mean(theta_mcmc > 0.6)                  # P(theta > 0.6 | y)
quantile(theta_mcmc, c(0.025, 0.975))   # 95% ETI
```

---

# In-Class Exercises

## Exercise 1: Posterior Summaries (Dutch Railways)

> Posterior: $\lambda \mid y \sim \text{Gamma}(8, 4.5)$.

::: {.panel-tabset}

### Exercise

1. Compute the posterior mean, mode, and median analytically.
2. Construct the 90%, 95%, and 99% ETI.
3. Compute $P(\lambda < 1.5 \mid y)$.
4. Compute $P(\lambda \leq 2 \mid y)$ and interpret.

### Solution

```{r ex1-solution}
s <- 8; r <- 4.5

cat("Mean:  ", round(s/r, 3),       "\n")
cat("Mode:  ", round((s-1)/r, 3),   "\n")
cat("Median:", round(qgamma(0.5, s, r), 3), "\n")

for (level in c(0.90, 0.95, 0.99)) {
  a <- (1 - level) / 2
  ci <- qgamma(c(a, 1-a), s, r)
  cat(level*100, "% ETI: [", round(ci[1],3), ",", round(ci[2],3), "]\n")
}

cat("P(lambda < 1.5 | y):", round(pgamma(1.5, s, r), 3), "\n")
cat("P(lambda <= 2  | y):", round(pgamma(2.0, s, r), 3), "\n")
```

**Interpretation:** About 68% posterior probability that the rate is at or below the safety threshold of 2/month.

:::

## Exercise 2: Interval Hypothesis Testing

> A researcher tests whether more than half of psychology studies replicate. Prior: Beta(4, 6). Observed: 58 of 100 studies replicate.

::: {.panel-tabset}

### Exercise

1. Derive the posterior.
2. Compute $BF_{+0}$ for $H_+: \theta > 0.5$ vs. $H_0: \theta \leq 0.5$.
3. Interpret the BF using the Kass & Raftery scale.
4. How much did beliefs change? Compare $P(\theta > 0.5)$ to $P(\theta > 0.5 \mid y)$.

### Solution

```{r ex2-solution}
# 1. Posterior: Beta(4+58, 6+42) = Beta(62, 48)
a_post <- 62; b_post <- 48

# 2. BF
prior_prob <- pbeta(0.5, 4, 6,     lower.tail = FALSE)
post_prob  <- pbeta(0.5, a_post, b_post, lower.tail = FALSE)
BF_plus0   <- (post_prob / (1 - post_prob)) / (prior_prob / (1 - prior_prob))

cat("Prior P(theta > 0.5):     ", round(prior_prob, 3), "\n")
cat("Posterior P(theta > 0.5 | y):", round(post_prob, 3), "\n")
cat("BF(+, 0) =", round(BF_plus0, 2), "\n")

plot_beta_binomial(alpha = 4, beta = 6, y = 58, n = 100)
```

:::

## Exercise 3: Savage–Dickey

> Same replication study. Test the point null $H_0: \theta = 0.5$.

::: {.panel-tabset}

### Exercise

1. Compute $BF_{01}$ using the Savage–Dickey formula.
2. Interpret $BF_{10}$.
3. Does this agree directionally with Exercise 2? Why might the magnitude differ?

### Solution

```{r ex3-solution}
theta0     <- 0.5
prior_dens <- dbeta(theta0, 4,     6    )
post_dens  <- dbeta(theta0, 62,    48   )

BF_01 <- post_dens / prior_dens
cat("BF_01 =", round(BF_01, 3), "\n")
cat("BF_10 =", round(1/BF_01, 3), "\n")
```

**Note:** The interval BF (Exercise 2) and the Savage–Dickey BF answer subtly different questions — one asks about a region, the other about a specific value. Both should agree *directionally* if the data are consistent.

:::

## Exercise 4: Grid Approximation with a Truncated Prior

> A researcher believes the placebo effect rate is between 30% and 90%, and is otherwise uncertain. Prior: Beta(2,2) truncated to $[0.30, 0.90]$. Observed: 18 of 25 patients improve.

::: {.panel-tabset}

### Exercise

1. Implement the grid approximation.
2. Compute the posterior mean and 95% ETI.
3. Compare to an untruncated Beta(2,2) — how much does the truncation matter here?

### Solution

```{r ex4-solution}
theta_grid <- seq(0, 1, length.out = 5000)
y_obs <- 18; n_obs <- 25

# Truncated prior
prior_trunc <- dbeta(theta_grid, 2, 2) *
               (theta_grid >= 0.3 & theta_grid <= 0.9)
lik         <- dbinom(y_obs, n_obs, prob = theta_grid)
post_trunc  <- (prior_trunc * lik) / sum(prior_trunc * lik)
samp_trunc  <- sample(theta_grid, 1e4, replace = TRUE, prob = post_trunc)

# Untruncated
prior_full <- dbeta(theta_grid, 2, 2)
post_full  <- (prior_full * lik) / sum(prior_full * lik)
samp_full  <- sample(theta_grid, 1e4, replace = TRUE, prob = post_full)

cat("Truncated: mean =", round(mean(samp_trunc), 3),
    " 95% ETI = [", round(quantile(samp_trunc, 0.025), 3),
    ",",             round(quantile(samp_trunc, 0.975), 3), "]\n")
cat("Full:      mean =", round(mean(samp_full),  3),
    " 95% ETI = [", round(quantile(samp_full,  0.025), 3),
    ",",             round(quantile(samp_full,  0.975), 3), "]\n")

df_plot <- data.frame(
  theta = c(samp_trunc, samp_full),
  Prior = rep(c("Truncated Beta(2,2)", "Full Beta(2,2)"), each = 1e4)
)
ggplot(df_plot, aes(x = theta, fill = Prior)) +
  geom_density(alpha = 0.5) +
  labs(title = "Posterior: truncated vs full prior",
       x = expression(theta), y = "Density") +
  theme_minimal(base_size = 13) +
  theme(legend.title = element_blank())
```

:::

---

<div class="note">
**Next lecture (June 15):** Posterior prediction — the posterior predictive distribution, prior predictive checks, model checking, and how hypothesis specification shapes the Bayes factor (Ch. 8 + Etz et al. 2018).
</div>
