---
title: "9. Linear Model (Regression)"
subtitle: "One predictor"
author: "Johnny van Doorn"
date: "18 September 2026"
view: "https://teams.microsoft.com/meet/330874021866475?p=QMChUAKXBz9m28zZWL"
footer: "Scientific & Statistical Reasoning"
format:
  revealjs:
    output-ext: slide.html
    css: ../../ajohnson.css
  html: default
css: ../../ajohnson.css
---

<div class="goals">
In this lecture we discuss:

* Linear model
  * Theory
  * Assumptions
  * In JASP
* Predictions and model fit
* Wrapping up: how many ways do we have for assessing an association?
    
<strong>Reading:</strong>
Chapter 8 (§8.1–8.8)
</div>


```{r child="regression_one_predictor.qmd"}
```


# Closing

## Next Week

- Exam!
  - Mix of using JASP/interpreting output/conceptual understanding
  - Field book available (like in this weeks WA), including glossary
  - WA will contain old exam questions

## Recommended Exercises

- [Exercise 8.1](https://discoverjasp.com/pages/exercises#chapter-8)

```{r child="../../footer.qmd"}
```
