---
title: "9. Linear Model (Regression)"
subtitle: "One predictor"
author: "Johnny van Doorn"
date: "18 September 2026"
view: "https://teams.microsoft.com/meet/330874021866475?p=QMChUAKXBz9m28zZWL"
footer: "Scientific & Statistical Reasoning"
format:
  revealjs:
    output-ext: slide.html
    css: ../../ajohnson.css
  html: default
css: ../../ajohnson.css
---

<div class="goals">
In this lecture we discuss:

* Linear model
  * Theory
  * Assumptions
  * In JASP
* Predictions and model fit

<strong>Reading:</strong>
Chapter 8 (§8.1–8.8)
</div>


```{r child="correlation-recap.qmd"}
```


```{r child="regression_one_predictor.qmd"}
```


# Closing

## Exam note

Alternative hypothesis:

- The question context tells you which test/alternative hypothesis to use
- Linear regression in JASP only reports two-sided $p$, so those stay two-sided

Assumptions:

- Only assessing the assumptions, not yet how to handle violations
- Only Sensitivity (scatterplot, casewise diagnostics), Normality (Q-Q plot), Linearity (Q-Q plot, scatterplot)
- Don't need to study ch6; lecture and application in ch8 suffice

See trial exam questions for examples of both


## Next Week

- Exam!
  - Mix of using JASP/interpreting output/conceptual understanding
  - WA will have trial exam with old exam questions
  - Field book available (like in the trial exam), including glossary


## Recommended Exercises

- [Exercise 8.1](https://discoverjasp.com/pages/smart_alex#task-8.1)

```{r child="../../footer.qmd"}
```
