# Last week {.section}

```{r recap-where-we-left-off, echo=FALSE}
source("../../_setup.R")
# Today's data, read here so the recap can already use it. regression_one_predictor.qmd
# reads it again for itself, so neither child depends on the other's variables.
album   <- read.csv(datasetsPath("2026/album_sales.csv"))
adverts <- album$Adverts
sales   <- album$Sales
```

## Last week: correlation & NHST {.smaller}
* $COV_{xy}$: association between two variables, in the original measurement units
* $r$:  completely standardized: $r_{xy} = \frac{COV_{xy}}{s_x s_y}$
* Get $p$ by converting to $t$: $t_r = \frac{r \sqrt{N-2}}{\sqrt{1 - r^2}}$
  * Use [the sampling distribution](https://statisticalreasoning-uva.shinyapps.io/NHST_Continuous/)
* Partial correlation: the same association, holding a third variable constant




## Two different units {.smaller}

```{r recap-two-different-units, echo=FALSE, fig.width=8, fig.height=4.8}
# Same build-up as the z-score slides in lecture 6, on this week's data: the
# first 20 albums, so individual points stay readable.
nAlb <- 20
adv  <- adverts[1:nAlb]
sal  <- sales[1:nAlb]

par(cex = 1.0, cex.lab = 1.1)
plot(adv, bg = 'orange', pch = 21, bty = "n", ylim = c(0, 1800), las = 1,
     ylab = "Variables", xlab = "Album nr.")
points(sal, bg = 'blue', pch = 21)
legend("topright", legend = c("Adverts (x £1,000)", "Sales (x 1,000)"),
       fill = c("orange", "blue"), bty = "n", cex = 0.8)
lines(rep(mean(adv), nAlb), col = 'orange')
lines(rep(mean(sal), nAlb), col = 'blue')
```

A deviation of 100 means something different in each variable

## Standardize {.smaller}

$$z_i = \frac{x_i - \bar{x}}{sd_x}$$

```{r recap-standardize, echo=FALSE, fig.width=8, fig.height=4.4}
z.adv <- (adv - mean(adv)) / sd(adv)
z.sal <- (sal - mean(sal)) / sd(sal)

par(cex = 1.0, cex.lab = 1.1)
plot(z.adv, bg = 'orange', pch = 21, bty = "n", ylim = c(-2.5, 2.5), las = 1,
     ylab = "z-scores", xlab = "Album nr.")
points(z.sal, pch = 21, bg = 'blue')
legend("topright", legend = c("z-adverts", "z-sales"),
       fill = c("orange", "blue"), bty = "n", cex = 0.8)

lines(rep(mean(z.adv), nAlb), col = 'orange')
lines(rep(mean(z.sal), nAlb), col = 'blue', lty = 2)

segments(1:nAlb, mean(z.adv), 1:nAlb, z.adv, col = 'orange')
segments(1:nAlb, mean(z.sal), 1:nAlb, z.sal, col = 'blue', lty = 2)
```

Mean = 0 and sd = 1, now the deviations are comparable.

## Correlation is the covariance of z-scores {.smaller}

$$z_i = \frac{x_i - \bar{x}}{s_x} \qquad\qquad r_{xy} = \frac{COV_{xy}}{s_x s_y} = COV_{z_x z_y}$$

```{r recap-correlation-is-the-covariance-of-z-scores, echo=TRUE}
z.adverts <- (adverts - mean(adverts)) / sd(adverts)
z.sales   <- (sales   - mean(sales))   / sd(sales)

cov(z.adverts, z.sales)

cor(adverts, sales)
```

If we standardize both variables first, the covariance *is* the correlation.

## Same covariance, three denominators {.smaller}

```{r recap-same-covariance-three-denominators, echo=TRUE}
cov(adverts, sales)                               # original scale

cov(adverts, sales) / (sd(adverts) * sd(sales))   # standardize x and y -> the correlation

cov(adverts, sales) / var(adverts)                # only x -> today's slope
```

Standardizing makes a statistic universally interpretable, but you lose information

## Note on one-sided vs. two-sided testing {.smaller}

A choice in the analysis before seeing the data (i.e., informed by the research question, not the result).

* Two-sided
  * Safer choice if unsure about the direction
  * "Spending $\alpha$" on two tails needs a more extreme test statistic to reach significance

* One-sided
  * Same $\alpha$, but the $\alpha \%$ rejection area is in 1 tail
  * Higher power, relevant when $N$ is small (expensive studies, rare cases)
  * Blind to an effect in the opposite direction

[Additional rambling here](https://johnnydoorn.github.io/teaching-statistics/extra-texts/one-sided-vs-two-sided.html)


## Today {.smaller}

* The regression slope $b_1$: $b_1 = r_{xy} \frac{s_y}{s_x}$
* Same trick for significance: coefficient $b$ $\rightarrow t \rightarrow p$
* Making predictions based on a model
* One predictor today, multiple predictors in block 3

---


