# Nonparametric tests {.section}

```{r}
#| echo: false
library("knitr")
```

- When assumptions are violated
    - e.g., strong non-normality
- When the variable is ordinal
    - e.g., when playing Mario Kart
- When unsure about outliers
    - e.g., some observations with z $\approx$ 3

+ [Topic of my dissertation!](https://dare.uva.nl/search?identifier=fca7cec7-5ded-4e99-8c79-a0d0ee6a3f4c)

## Parametric vs Nonparametric {#mytable}

<!-- <style>
#mytable th {
background: -webkit-gradient(linear, 50% 0%, 50% 100%, color-stop(40%, #FFA500), color-stop(80%, #FFA500)) no-repeat !important;
background-image: -webkit-linear-gradient(top, rgb(255,165,0) 40%, rgb(255,165,0) 80%) !important;
background: -webkit-linear-gradient(top, #FFA500 40%, #FFA500 80%) no-repeat !important;
background-image: -webkit-linear-gradient(top, rgb(255,165,0) 40%, rgb(255,165,0) 80%) !important;
background: -moz-linear-gradient(top, #FFA500 40%, #FFA500 80%) no-repeat !important;
background: -o-linear-gradient(top, #FFA500 40%, #FFA500 80%) no-repeat !important;
background: linear-gradient(top, #FFA500 40%, #FFA500 80%) no-repeat !important;
}
</style> -->

Attribute    | Parametric            | Nonparametric
-------------|-----------------------|--------------------------
distribution | normally distributed  | any distribution
sampling     | random sample         | random sample
sensitivity to outliers | yes        | no
works with   | large data sets       | small and large data sets
speed        | fast                  | slow

## Ranking {.subsection}

```{r, fig.height=3, fig.align='center'}
x <- c(1, 4, 6, 7, 8, 9)
y <- c(1, 4, 6, 7, 8, 39)

layout(matrix(1:2, 1, 2))
boxplot(x, horizontal=T, col='red')
boxplot(y, horizontal=T, col='red')
```

```{r, fig.height=3}
kable(rbind(rx = rank(x), ry = rank(y)))
```

## Ties {.smaller}
```{r, echo=TRUE}
# Scores
x <- c(11, 42, 62, 73, 84, 84, 42, 73, 84)

# Sort x
x <- sort(x)

# Assign ranks
ranks <- rank(x)
```

```{r, echo=FALSE}
kable(rbind(index = 1:length(x), x, ranks))
```


$$\frac{2 + 3}{2} = 2.5,  \frac{5 + 6}{2} = 5.5,  \frac{7 + 8 + 9}{3} = 8$$

## Procedure {.smaller}

1. Assumption: independent random samples.
2. Hypothesis: 
<br>$H_0$ : equal population distributions (implies equal mean ranking) 
<br>$H_A$ : unequal mean ranking (two sided) 
<br>$H_A$ : higher mean ranking for one group.
3. Test statistic is difference between mean or sum of ranking.
4. Standardize test statistic to normal sampling distribution.
5. Calculate *P*-value one or two sided.
5. Conclude to reject $H_0$ if $p < \alpha$.