---
title: "11. T-tests"
author: "Johnny van Doorn"
date: "30 sep 2025"
footer: "Scientific & Statistical Reasoning"
format:
  revealjs:
    output-ext: slide.html
    css: ../../ajohnson.css
  html: default
css: ../../ajohnson.css
---

# Block 2: Life is Mean

- Comparing 2 means (t-test)
- Comparing 2 or more means (ANOVA)
- Between subjects vs. Within subjects
- Adding predictors
    
--- 

<div class="goals">
In this lecture we aim to:

* Introduce the t-test for comparing two means
  * Between subjects
  * Within subjects
* Regression?!
    
<strong>Reading:</strong>
Chapters 9
</div>

```{r child="../../topics/t-test_one-sample/One-sample_t-test.qmd"}
```


```{r child="../../topics/t-test_paired/t-test_paired.qmd"}
```

```{r child="../../topics/t-test_independent/t-test_independent.qmd"}
```

# Closing

## Recommended Exercises
::: {}
- [Exercise 5.1](https://discoverjasp.com/pages/smart_alex#task-5.1), [Exercise 5.5](https://discoverjasp.com/pages/smart_alex#task-5.5)

- [Exercise 7.1](https://discoverjasp.com/pages/smart_alex#task-7.1), [Exercise 7.6](https://discoverjasp.com/pages/smart_alex#task-7.6), [Exercise 7.7](https://discoverjasp.com/pages/smart_alex#task-7.7)
    - Note: `pets.jasp` can also be downloaded from the [Data page](https://discoverjasp.com/pages/data)

:::


```{r child="../../footer.qmd"}
```
