---
title: "5. JASP: Visualization and Correlation"
author: "Johnny van Doorn"
date: "10 sep 2025"
footer: "Scientific & Statistical Reasoning"
format:
  revealjs:
    output-ext: slide.html
    css: ../../ajohnson.css
  html: default
css: ../../ajohnson.css
---

<div class="goals">
In block 1 we aim to:

* Discuss NHST concepts
* Introduce JASP
* Correlation
* Linear model (regression)
    
<strong>Reading:</strong>
Chapters 1-8.8
</div>

--- 

<div class="goals">
In this lecture we aim to:

* Introduce JASP
  * Descriptive statistics
  * Data visualizations
* Correlation
  * How it works
  * Controlling for a third variable
    
<strong>Reading:</strong>
Chapters 4, 5, 7
</div>


```{r child="../../topics/JASP/JASP.qmd", eval=TRUE}
```


```{r child="../../topics/visualization/visualization.qmd", eval=TRUE}
```

```{r child="../../topics/correlation/correlation.qmd", eval=TRUE}
```

## Next Time

- The linear model
  - Predictions
  - Assumptions
  - Fun

## Bored? 
::: {}
- [Exercise 5.1](https://discoverjasp.com/pages/smart_alex#task-5.1), [Exercise 5.5](https://discoverjasp.com/pages/smart_alex#task-5.5)

- [Exercise 7.1](https://discoverjasp.com/pages/smart_alex#task-7.1), [Exercise 7.6](https://discoverjasp.com/pages/smart_alex#task-7.6), [Exercise 7.7](https://discoverjasp.com/pages/smart_alex#task-7.7)
    - Note: `pets.jasp` can also be downloaded from the [Data page](https://discoverjasp.com/pages/data)

:::

<!-- Footer insert below -->

```{r child="../../footer.qmd"}
```
