---
title: "19. Chi-squared test"
subtitle: "Contingency Tables"
author: "Johnny van Doorn"
date: "16 October 2026"
footer: "Scientific & Statistical Reasoning"
format:
  revealjs:
    output-ext: slide.html
    css: ../../ajohnson.css
  html: default
css: ../../ajohnson.css
---

<div class="goals">
In this lecture we aim to:

- Discuss association between categorical variables
  - Contingency tables
- Show this test in JASP
    
<strong>Reading:</strong>
Chapter 16 (§16.1–16.10)
</div>


```{r child="chi_square_test.qmd"}
```


## JASP

![](../../images/JASP_logo_green.svg){width="300px"}


# Closing

## Recap

- Chi-square test assesses categorical associations
- Can be used to assess (in)dependence between predictors
- Has test statistic (χ²) + effect size (odds ratio)

## Next Time

- Non-parametric tests: rank-based equivalents of old friends (lecture 20)

## Recommended Exercises

- [Exercise 16.1](https://discoverjasp.com/pages/smart_alex#task-16.1)
- [Exercise 16.6](https://discoverjasp.com/pages/smart_alex#task-16.6)
- [Exercise 16.7](https://discoverjasp.com/pages/smart_alex#task-16.7)

```{r child="labcoat-leni.qmd"}
```

```{r child="../../footer.qmd"}
```
