---
title: "18. RM ANOVA"
subtitle: "Repeated & Mixed"
author: "Johnny van Doorn"
date: "TODO 2026"
footer: "Scientific & Statistical Reasoning"
format:
  revealjs:
    output-ext: slide.html
    css: ../../ajohnson.css
  html: default
css: ../../ajohnson.css
---

<div class="goals">
In this lecture we aim to:

- Refresh some concepts from previous lectures
- Introduce RM ANOVA
- Introduce Mixed ANOVA
- Show these in JASP
    
<strong>Reading:</strong>
Chapter 14 (§14.1–14.18)
</div>


```{r child="ANOVA_repeated_one-way.qmd"}
```


```{r child="ANOVA_repeated_factorial.qmd"}
```


```{r child="ANOVA_mixed.qmd"}
```


# Closing

## Recap

- In a repeated measures design we can account for baseline differences, to reduce the overall model error
- With within-subjects predictors, we need to check:
  - Sphericity (equal variances of difference scores)
- With between-subjects predictors, we need to check:
  - Equal variances of groups

## Recommended Exercises

- [Exercise 14.4](https://discoverjasp.com/pages/exercises#chapter-14)
- [Exercise 14.5](https://discoverjasp.com/pages/exercises#chapter-14)
- To understand mechanics of RM ANOVA, [Exercise 14.1](https://discoverjasp.com/pages/exercises#chapter-14)
- No sums of squares or effect size calculations required for the exam

```{r child="../../footer.qmd"}
```
