---
title: "20. Non-parametric tests"
author: "Johnny van Doorn"
date: "21 oct 2025"
footer: "Scientific & Statistical Reasoning"
format:
  revealjs:
    output-ext: slide.html
    css: ../../ajohnson.css
  html: default
css: ../../ajohnson.css
---

<div class="goals">
In this lecture we aim to:

- Repeat some assumptions
- Introduce rank-based methods
- Look at rank-based equivalents to some old friends
- Show these in JASP
    
<strong>Reading:</strong>
Chapter 15
</div>

## Assumptions
- Not an all-or-nothing case
- Assess the degree of the violation
- Especially vulnerable with low sample sizes
- Visualize (Q-Q plot, boxplots, descriptives)
- Sometimes there are cures, sometimes not...
- Don't trash your data, but investigate: why is this occurring?

## Homogeneity of variances
- Between subjects designs
- Correction: Welch for t-test/one-way ANOVA
- Adjusts df $\rightarrow p$-value & CI
- Jane Superbrain 14.3 (pointless p)

## Sphericity
- Within subjects designs with >2 levels
- Correction: Greenhouse-Geisser for RM ANOVA
- Adjusts df $\rightarrow p$-value & CI
- Jane Superbrain 14.3 (pointless p)

<!-- ## Overlap (association) between predictors -->

## Normality
- Shape of distribution (and no CLT)
- Outliers
- [Illustration of consequences](https://johnnydoorn.github.io/teaching-statistics/extra-texts/ViolatingNormalityANOVA.html)


```{r child="../../topics/non_parametric_concept/non_parametric_concept.qmd", eval=TRUE}
```

```{r child="../../topics/nonpar_wilcoxon_rank_sum_test/wilcoxon_rank_sum_test.qmd", eval=TRUE}
```

```{r child="../../topics/nonpar_wilcoxon_signed_rank_test/wilcoxon_signed_rank_test.qmd", eval=TRUE}
```

```{r child="../../topics/nonpar_kruskal_wallis_test/kruskal_wallis_test.qmd", eval=TRUE}
```

```{r child="../../topics/nonpar_friedman_anova/friedman_anova.qmd", eval=TRUE}
```

# Closing

## Recap

- Non-parametric tests are rank-based, which means they are robust to:
  - Non-normality
  - Transformations of the data
  - Outliers
  
## Recommended Exercises

::: {}
- [Exercise 15.1](https://discoverjasp.com/pages/smart_alex#task-15.1),  [Exercise 15.2](https://discoverjasp.com/pages/smart_alex#task-15.2), [Exercise 15.4](https://discoverjasp.com/pages/smart_alex#task-15.4), [Exercise 15.11](https://discoverjasp.com/pages/smart_alex#task-15.11)
:::

```{r child="../../footer.qmd"}
```

