---
title: "22. Multiple regression"
author: "Johnny van Doorn"
date: "29 oct 2025"
footer: "Scientific & Statistical Reasoning"
format:
  revealjs:
    output-ext: slide.html
    css: ../../ajohnson.css
  html: default
css: ../../ajohnson.css
---

# Block 3: Life is Continuous

- Regression with multiple predictors
- Interaction effects
- Mediation effects
- Chi-square
- Bayes!
    
--- 

<div class="goals">
In this lecture we aim to:

- Repeat some regression
- Look at multiple predictors
- Show these in JASP
    
<strong>Reading:</strong>
Chapter 8
</div>


<!-- ## WA 10 Data -->

<!-- - [Salary (partial correlation)](https://johnnydoorn.github.io/teaching-statistics/datasets/Salary.jasp) -->
<!-- - [Health (regression)](https://johnnydoorn.github.io/teaching-statistics/datasets/Health.jasp) -->

```{r child="../../topics/regression_multiple_predictors/regression_multiple_predictors.qmd"}
```

# Closing

## Recap

- In linear regression, we can build multiple models and look at their fit
- Just as in ANOVA, when we have multiple predictors we look out for:
  - Dependence between predictors
  - Interaction effects

  
## Recommended Exercises

::: {}
- [Exercise 8.4](https://discoverjasp.com/pages/smart_alex#task-8.4),  [Exercise 8.7](https://discoverjasp.com/pages/smart_alex#task-8.7)
:::

```{r child="../../footer.qmd"}
```

