---
title: "23. Mediation"
author: "Johnny van Doorn"
date: "30 October 2026"
footer: "Scientific & Statistical Reasoning"
format:
  revealjs:
    output-ext: slide.html
    css: ../../ajohnson.css
  html: default
css: ../../ajohnson.css
---

<div class="goals">
In this lecture we aim to:

- Explore mediation analysis
- Link it to regression
- Demonstrate in JASP
    
<strong>Reading:</strong>
Chapter 10 (§10.4–10.7)
</div>


```{r child="mediationNames.qmd"}
```


## JASP

![](../../images/JASP_logo_green.svg){width="300px"}


# Closing

## Recap

- Mediation analysis is a way to model dependence between predictors
- PROCESS module is very flexible
- Causal assumptions extremely important with PROCESS models

## Next Time

- Moderation: interactions between continuous predictors (lecture 26)

## Recommended Exercises

- [Exercise 10.4](https://discoverjasp.com/pages/smart_alex#task-10.4)
- [Exercise 10.5](https://discoverjasp.com/pages/smart_alex#task-10.5)

```{r child="labcoat-leni.qmd"}
```

```{r child="../../footer.qmd"}
```
