---
title: "15. Factorial ANOVA"
subtitle: "Independent factorial"
author: "Johnny van Doorn"
date: "08 oct 2025"
footer: "Scientific & Statistical Reasoning"
format:
  revealjs:
    output-ext: slide.html
    css: ../../ajohnson.css
  html: default
css: ../../ajohnson.css
---

<div class="goals">
In this lecture we aim to:

- ANOVA with 2 categorical predictors
- Discuss interaction
- Demonstration in JASP
    
<strong>Reading:</strong>
Chapter 13
</div>


```{r child="../../topics/ANOVA_independent_factorial/factorial_independent_ANOVA.qmd"}
```


# Closing

## Recap

With multiple predictor variables:

- We are still explaining variance
- We check the interaction
  - Plots!
  - F-value of interaction effect
  - Conditional post-hoc tests

## Recommended Exercises

::: {}
- [Exercise 13.1](https://discoverjasp.com/pages/smart_alex#task-13.1), [Exercise 13.3](https://discoverjasp.com/pages/smart_alex#task-13.3),  [Exercise 13.6](https://discoverjasp.com/pages/smart_alex#task-13.6)
- You do not need to compute effects sizes by hand!
:::

```{r child="../../footer.qmd"}
```
