# Independent-samples t-test

Compare 2 independent samples

## Independent-samples t-test

In the independent-samples t-test the mean of both independent samples is calculated and the difference of these $(\bar{X}_1 - \bar{X}_2)$ means is tested against the null hypothesis where $\mu = 0$.

$$t_{n_1 + n_2 -2} = \frac{(\bar{X}_1 - \bar{X}_2) - \mu}{{SE}_p}$$
Where $n_1$ and $n_2$ are the number of cases in each group and $SE_p$ is the pooled standard error.

## Hypothesis

\begin{aligned}
H_0 &: t = 0 = \mu_t \\
H_A &: t \neq 0 \\
H_A &: t > 0 \\
H_A &: t < 0 \\
\end{aligned}

## Data structure

index | k | outcome
------|----|--
1     | 1 | x
2     | 1 | x
3     | 2 | x
4     | 2 | x

Where $k$ is the level of the categorical predictor variable and $x$ is the value of the outcome/dependent variable.

## Additional assumption

Specific for independent sample $t$-test.

* Equality of variance
    * $H_0$ : Variance $=$ equal ($p$ > .05)
    * $H_A$ : Variance $\neq$ equal ($p$ < .05)

## Example

We are going to use the IQ estimates we collected last week again. You had to guess the IQ of the one sitting next to you and your own IQ. But we are going to add class position to the data set. We did not register this so we are going to simulate some positions.

```{r, warning=FALSE, echo=FALSE, message=FALSE}
if(!'gsheet' %in% installed.packages()) { install.packages('gsheet') }
library("gsheet")
data <- gsheet2tbl('https://docs.google.com/spreadsheets/d/1LN8j2MgTk1Su-hZCbFjexiUKQx-LmbgGqlpN6YzF2jQ/edit?usp=sharing')[, -1]

# data = data[grep("2017", as.character(data$Timestamp)),c("IQ next to you", "Own IQ")]

nextIQ <- data$nextIQ
ownIQ  <- data$ownIQ
```


```{r, echo=TRUE}
set.seed(23)
classPosition <- sample(c("front", "back"), dim(data)[1], replace = TRUE)
```

## The data

```{r, echo=FALSE}
data$classPosition <- classPosition

# Write data for use in SPSS
write.table(data, "IQ.csv", row.names=FALSE, col.names=TRUE, dec=',')

DT::datatable(data[, c('classPosition', 'ownIQ')], options = list(searching = FALSE, scrollY = 415, paging = F, info = F))
```

## Calculate means {.subsection}

```{r, echo=TRUE}
ownIQfront  <- subset(data, classPosition == "front",   select = ownIQ)$ownIQ
ownIQback <- subset(data, classPosition == "back", select = ownIQ)$ownIQ

ownIQfront.mean <- mean(ownIQfront, na.rm = TRUE)
ownIQback.mean <- mean(ownIQback, na.rm = TRUE)

rbind(ownIQfront.mean, ownIQback.mean)
```

## Calculate variance {.smaller .subsection}

```{r, echo=TRUE}
ownIQfront.var   <- var(ownIQfront,   na.rm = TRUE)
ownIQback.var <- var(ownIQback, na.rm = TRUE)
print(rbind(ownIQfront.var, ownIQback.var))

ownIQfront.n   <- length(ownIQfront)   - 1
ownIQback.n <- length(ownIQback) - 1
print(rbind(ownIQfront.n, ownIQback.n))
```


## Calculate t-value {.smaller}

$$t_{n_1 + n_2 -2} = \frac{(\bar{X}_1 - \bar{X}_2) - \mu}{{SE}_p}$$

Where ${SE}_p$ is the pooled standard error.

$${SE}_p = \sqrt{\frac{S^2_p}{n_1}+\frac{S^2_p}{n_2}}$$

And $S^2_p$ is the pooled variance.

$$S^2_p = \frac{(n_1-1)s^2_1+(n_2-1)s^2_2}{n_1+n_2-2}$$

Where $s^2$ is the variance and $n$ the sample size.

## Calculate pooled variance {.subsection}

$$S^2_p = \frac{(n_1-1)s^2_1+(n_2-1)s^2_2}{n_1+n_2-2}$$

```{r, echo=TRUE}
df <- ownIQfront.n + ownIQback.n - 2
pooledVar <- ( (ownIQfront.n-1)*ownIQfront.var + (ownIQback.n-1)*ownIQback.var ) / df

df
pooledVar
```

## Calculate pooled SE {.subsection}

$$ {SE}_p = \sqrt{\frac{S^2_p}{n_1}+\frac{S^2_p}{n_2}} $$

```{r, echo=TRUE}
sePooled <- sqrt( ((pooledVar/ownIQfront.n) + (pooledVar/ownIQback.n)) )
sePooled
```

## Calculate t-value {.subsection}

$$t_{n_1 + n_2 -2} = \frac{(\bar{X}_1 - \bar{X}_2) - \mu}{{SE}_p}$$

```{r, echo=TRUE}
tStat <- ( ownIQfront.mean - ownIQback.mean ) / sePooled

tStat
```

## Test for significance

Two tailed

```{r, echo=TRUE}
#| output-location: slide
if(!"visualize" %in% installed.packages()) { install.packages("visualize") }
library("visualize")

if (df == 2) df <- df + 1 # cheat to be able to build slides.. 
visualize.t(c(-round(tStat, 2), round(tStat, 2)), df+1, section="tails")
```

## Effect-size

$$r = \sqrt{\frac{t^2}{t^2 + \text{df}}}$$

```{r, echo=TRUE}
r <- sqrt(tStat^2/(tStat^2 + df))

r

d <- 2*tStat / sqrt(df)
d
```


## But what about equal variances?!

There exist different hypothesis tests for this - the most used is Levene's test:
```{r, echo=TRUE, warning=FALSE, message=FALSE}

if(!"car" %in% installed.packages()) { install.packages("car") }
library("car")

leveneTest(y = ownIQ, group = classPosition)
```

## The difference only matters when $n_A \neq n_B$!


```{r, echo=TRUE, warning=FALSE, message=FALSE}
meanA <- 50
meanB <- 55

sdA <- 5
sdB <- 10

nA <- 20
nB <- 20

seA <- sdA / sqrt(nA)    
seB <- sdB / sqrt(nB)  
tStatWelch <- (meanA - meanB) / sqrt((sdA^2 / nA) + (sdB^2 / nB))


pooledS <- sqrt(((nA - 1) * sdA^2 + (nA - 1) * sdB^2) / (nA + nB - 2))
tStatStudent <- (meanA - meanB) / (pooledS *  sqrt(1/nA + 1/nA))

tStatStudent
tStatWelch
```