---
title: "19. Bayesian ANOVA"
subtitle: "Parameter estimation & hypothesis testing"
author: "Johnny van Doorn"
date: "22 oct 2024"
footer: "Scientific & Statistical Reasoning"
format:
  revealjs:
    output-ext: slide.html
    css: ../../ajohnson.css
  html: default
css: ../../ajohnson.css
---

<div class="goals">
In this lecture we aim to:

- Repeat some assumptions
- Introduce rank-based methods
- Look at rank-based equivalents to some old friends
- Show these in JASP
    
<strong>Reading:</strong>
Chapter 15

## Course literature

- [Repitition from RMS](https://johnnydoorn.github.io/IntroductionBayesianInference/)
- [Addition: chapter 7 on ANOVA](https://johnnydoorn.github.io/IntroductionBayesianInference/07-anova.html)

- [All data sets in Field](https://johnnydoorn.github.io/discoverjasp/repository/dsuj_sav_files.zip)


```{r child="../../topics/bayes/bayesian_parameter_estimation.qmd"}
```

```{r child="../../topics/bayes/bayesian_hypothesis_testing.qmd"}
```

```{r child="../../topics/bayes/bayesian_anova.qmd"}
```


# Closing

## Recap

- Non-parametric tests are rank-based, which means they are robust to:
  - Non-normality
  - Transformations of the data
  - Outliers
  
## Recommended Exercises

::: {}
- [Exercise 15.1](https://discoverjasp.com/pages/smart_alex#task-15.1),  [Exercise 15.2](https://discoverjasp.com/pages/smart_alex#task-15.2), [Exercise 15.4](https://discoverjasp.com/pages/smart_alex#task-15.4), [Exercise 15.11](https://discoverjasp.com/pages/smart_alex#task-15.11)
:::
<!-- Footer insert below -->

```{r child="../../footer.qmd"}
```
