---
title: "29. Bayesian ANOVA"
subtitle: "Parameter estimation & hypothesis testing"
author: "Johnny van Doorn"
date: "14 nov 2025"
footer: "Scientific & Statistical Reasoning"
format:
  revealjs:
    output-ext: slide.html
    css: ../../ajohnson.css
  html: default
css: ../../ajohnson.css
---

<div class="goals">
In this lecture we aim to:

- Repeat some core Bayesian concepts
- Apply these to ANOVA
- Show Bayesian ANOVA in JASP
    
<strong>Reading:</strong>
[A Brief Introduction to Bayesian Inference](https://johnnydoorn.github.io/IntroductionBayesianInference/07-anova.html)
</div>



```{r child="../../topics/bayes/bayesian_parameter_estimation.qmd"}
```

```{r child="../../topics/bayes/bayesian_hypothesis_testing.qmd"}
```

```{r child="../../topics/bayes/bayesian_anova.qmd"}
```


# Closing

## Recap

- Bayesian statistics is a thing
- The fun doesn't stop at t-test/correlation, but extends to ANOVA/regression
  - Obtain evidence for $H_0$
  - Gradual assessment of evidence
  - Need to specify prior distributions
  
## Recommended Exercises

::: {}
- Revisit Labcoat Leni's and do a Bayesian analysis
  - [Chapter 9 - T-test](https://discoverjasp.com/pages/labcoat_leni#chapter-9)
    - My take: [jasp](https://johnnydoorn.github.io/teaching-statistics/datasets/leni_09b-bayes.jasp) & [html](https://johnnydoorn.github.io/teaching-statistics/datasets/leni_09b-bayes.html)
  - [Chapter 11 - ANOVA](https://discoverjasp.com/pages/labcoat_leni#chapter-11)
    - My take: [jasp](https://johnnydoorn.github.io/teaching-statistics/datasets/leni_10-bayes.jasp) & [html](https://johnnydoorn.github.io/teaching-statistics/datasets/leni_10-bayes.html)
  - [Chapter 13 - Factorial ANOVA](https://discoverjasp.com/pages/labcoat_leni#chapter-13)
    - My take: [jasp](https://johnnydoorn.github.io/teaching-statistics/datasets/leni_13-bayes.jasp) & [html](https://johnnydoorn.github.io/teaching-statistics/datasets/leni_13-bayes.html)


:::
<!-- Footer insert below -->

```{r child="../../footer.qmd"}
```
