---
title: "20. Non-parametric tests"
author: "Johnny van Doorn"
date: "TODO 2026"
footer: "Scientific & Statistical Reasoning"
format:
  revealjs:
    output-ext: slide.html
    css: ../../ajohnson.css
  html: default
css: ../../ajohnson.css
---

<div class="goals">
In this lecture we aim to:

- Repeat some assumptions
- Introduce rank-based methods
- Look at rank-based equivalents to some old friends
- Show these in JASP
    
<strong>Reading:</strong>
Chapter 15 (§15.1–15.9)
</div>


```{r child="non_parametric_concept.qmd"}
```

```{r child="wilcoxon_rank_sum_test.qmd"}
```

```{r child="wilcoxon_signed_rank_test.qmd"}
```

```{r child="kruskal_wallis_test.qmd"}
```

```{r child="friedman_anova.qmd"}
```


# Closing

## Recap

- Non-parametric tests are rank-based, which means they are robust to:
  - Non-normality
  - Transformations of the data
  - Outliers

## Recommended Exercises

- [Exercise 15.1](https://discoverjasp.com/pages/exercises#chapter-15)
- [Exercise 15.2](https://discoverjasp.com/pages/exercises#chapter-15)
- [Exercise 15.4](https://discoverjasp.com/pages/exercises#chapter-15)
- [Exercise 15.11](https://discoverjasp.com/pages/exercises#chapter-15)

```{r child="../../footer.qmd"}
```
