# Mixed design ANOVA {.section}

```{r, echo=FALSE}
rm(list=ls())
if(!"DT" %in% installed.packages()) { install.packages("DT") }
library("DT")
```

## Mixed design 

The mixed ANOVA analyses the variance of the model while reducing the error by the within person variance.

* 1 dependent/outcome variable
* 2 or more independent/predictor variable with different subjects
    * 2 or more levels
* 1 or more independent/predictor variable with same subjects
    * 2 or more levels

## Assumptions

Same as repeated measures ANOVA and same as factorial ANOVA.

## Example {.smaller}

* Dependent variable
    * Accidents
* Independent variables
    * Speed (same subjects)
        * None
        * Small
        * Large
    * Alcohol (same subjects)
        * None
        * Small
        * Large
    * Daytime
        * Morning
        * Evening


## Data {.smaller}

```{r, echo=FALSE}
data <- read.csv(file = "RM_ANOVA_AlcoholSpeedTime.csv", header = TRUE)

# data[,-1] <- round(data[,-1],2)

datatable(data, 
          rownames = FALSE, 
          extensions = 'Buttons',
          options = list(searching = FALSE, 
                         scrollY = 415, 
                         paging = FALSE, 
                         info = FALSE,
                         dom = 'Bfrtip',
                         buttons   = c('csv'))
        )
```


