---
title: "13. ANCOVA"
author: "Johnny van Doorn"
date: "2 October 2026"
view: "https://amsuni-my.sharepoint.com/:v:/g/personal/j_b_vandoorn_uva_nl/IQA7LdI082EvRIBMc7O0Z4ghAc_R5Gt7pwDoVgncEjVHGhs?e=Q8B1Lz&nav=eyJyZWZlcnJhbEluZm8iOnsicmVmZXJyYWxBcHAiOiJTdHJlYW1XZWJBcHAiLCJyZWZlcnJhbFZpZXciOiJTaGFyZURpYWxvZy1MaW5rIiwicmVmZXJyYWxBcHBQbGF0Zm9ybSI6IldlYiIsInJlZmVycmFsTW9kZSI6InZpZXcifX0%3D"
footer: "Scientific & Statistical Reasoning"
format:
  revealjs:
    output-ext: slide.html
    css: ../../ajohnson.css
  html: default
css: ../../ajohnson.css
---

<div class="goals">
In this lecture we aim to:

- Introduce ANCOVA
- Implications of multiple predictors
- Show ANCOVA in JASP
    
<strong>Reading:</strong>
Chapter 12 (§12.1–12.12)
</div>


```{r child="ANCOVA-puppies.qmd"}
```

<!-- Backup: the simulated nationality/openness version. Swap it back in for ANCOVA-puppies.qmd above. -->
<!-- ```{r child="ANCOVA.qmd"} -->
<!-- ``` -->


# Closing

## Recap

- We can add a continuous predictor to the ANOVA model, to reduce the overall model error
- With multiple predictors, we need to check:
  - Whether there is an interaction effect
    - An existing interaction effect gives crucial context to main effects
  - Whether the predictors are associated
    - Associated predictors can either increase or reduce each other's unique contributions to explained variance

## Exam note: checking assumptions {.smaller}

What you should know: how to assess, robust alternatives, and why these assumptions matter

- No assumption NHST (Shapiro-Wilk, Levene's) on the exam (see Jane Superbrain Box 6.7)
- Normality: Q-Q plot of the residuals, with confidence interval
  - Assumption Checks → Q-Q plot residuals → Confidence interval
  - Points outside the band suggest non-normality
- Equal variances: ratio of the largest to the smallest variance (Section 6.9.6)
  - Ratio of 2 or more suggests unequal variances
  - Equivalently for sd's: ratio of $\sqrt{2} \approx 1.4$ or more

## Exam note: robust alternatives {.smaller}

- Unequal variances:
  - Welch $t$-test
  - Welch $F$ (one-way ANOVA *only*: Assumption Checks → Homogeneity corrections)
- Non-normality: nonparametric test (Chapter 15; lecture 20)
  - Mann-Whitney, Wilcoxon signed-rank, Kruskal-Wallis
- Question can give a decision rule, e.g., "use nonparametric test if more than 5 points fall outside the Q-Q confidence band"
- ANCOVA: no robust version, so have to eat the grain of salt if normality or variances seem problematic

## Exam note: ANCOVA assumptions {.smaller}

- Independence of the covariate and treatment effect
  - ANOVA with the covariate as dependent variable
  - What it means: groups differ on the covariate
- Homogeneity of regression slopes: interaction covariate * independent variable
  - Model → Model Terms: add or remove the interaction
  - What it means: effect of the covariate differs between groups

## Next Time

- Extensions of ANOVA:
  - Adding a second categorical variable (lecture 17)
  - Within-subjects measurements (lecture 18)
  - Mixed designs (lecture 18)

## Recommended Exercises

- [Exercise 12.2](https://discoverjasp.com/pages/smart_alex#task-12.2)
- [Exercise 12.5](https://discoverjasp.com/pages/smart_alex#task-12.5)
- [Exercise 12.6](https://discoverjasp.com/pages/smart_alex#task-12.6)

```{r child="labcoat-leni.qmd"}
```

```{r child="../../footer.qmd"}
```
