---
title: "13. ANCOVA"
author: "Johnny van Doorn"
date: "2 October 2026"
view: "https://amsuni-my.sharepoint.com/:v:/g/personal/j_b_vandoorn_uva_nl/IQA7LdI082EvRIBMc7O0Z4ghAc_R5Gt7pwDoVgncEjVHGhs?e=Q8B1Lz&nav=eyJyZWZlcnJhbEluZm8iOnsicmVmZXJyYWxBcHAiOiJTdHJlYW1XZWJBcHAiLCJyZWZlcnJhbFZpZXciOiJTaGFyZURpYWxvZy1MaW5rIiwicmVmZXJyYWxBcHBQbGF0Zm9ybSI6IldlYiIsInJlZmVycmFsTW9kZSI6InZpZXcifX0%3D"
footer: "Scientific & Statistical Reasoning"
format:
  revealjs:
    output-ext: slide.html
    css: ../../ajohnson.css
  html: default
css: ../../ajohnson.css
---

<div class="goals">
In this lecture we aim to:

- Introduce ANCOVA
- Implications of multiple predictors
- Show ANCOVA in JASP
    
<strong>Reading:</strong>
Chapter 12 (§12.1–12.12)
</div>


```{r child="ANCOVA-puppies.qmd"}
```

<!-- Backup: the simulated nationality/openness version. Swap it back in for ANCOVA-puppies.qmd above. -->
<!-- ```{r child="ANCOVA.qmd"} -->
<!-- ``` -->


# Closing

## Recap

- We can add a continuous predictor to the ANOVA model, to reduce the overall model error
- With multiple predictors, we need to check:
  - Whether there is an interaction effect
    - An existing interaction effect gives crucial context to main effects
  - Whether the predictors are associated
    - Associated predictors can either increase or reduce each other's unique contributions to explained variance

<!-- Also included in lecture 20's exam-note recap. -->
```{r child="exam-notes.qmd"}
```

## Next Time

- Extensions of ANOVA:
  - Adding a second categorical variable (lecture 17)
  - Within-subjects measurements (lecture 18)
  - Mixed designs (lecture 18)

## Recommended Exercises

- [Exercise 12.2](https://discoverjasp.com/pages/smart_alex#task-12.2)
- [Exercise 12.5](https://discoverjasp.com/pages/smart_alex#task-12.5)
- [Exercise 12.6](https://discoverjasp.com/pages/smart_alex#task-12.6)

```{r child="labcoat-leni.qmd"}
```

```{r child="../../footer.qmd"}
```
