---
title: "Scientific & Statistical Reasoning"
subtitle: "Statistics materials: slides, applets, datasets & extra explanations"
author: "Johnny van Doorn"
date: today
format:
  html:
    theme:
      light: [cosmo, overview-styles.scss]
      dark: [darkly, overview-styles.scss]
    toc: true
    toc-location: left
    toc-title: "On this page"
    title-block-banner: "#C34A5E"
    title-block-banner-color: white
    smooth-scroll: true
    anchor-sections: true
    link-external-newwindow: true
---

::: {.lead-intro}
This page is continuously updated with everything you need for the course: the lecture slides and handouts, interactive applets, the JASP datasets and result files, and extra written explanations. Use the toggle in the sidebar to switch between light and dark mode.
:::

## Slides

```{r child="../ssr/2026/ssr.qmd"}
```


## Applets
```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(stringr)
library("lubridate")
library("dplyr")
library('kableExtra')
allLinks <- c(
   "https://statisticalreasoning-uva.shinyapps.io/NHST_Binomial",
   "https://statisticalreasoning-uva.shinyapps.io/NHST_Continuous",
   "https://statisticalreasoning-uva-2.shinyapps.io/SimpleRegression"
  # "https://statisticalreasoning-uva.shinyapps.io/NHST_Fdistribution",
  # "https://statisticalreasoning-uva.shinyapps.io/onewayANOVA",
  # "https://statisticalreasoning-uva.shinyapps.io/ANCOVA",
  # "https://statisticalreasoning-uva-2.shinyapps.io/Multi-ANOVA",
  # "https://statisticalreasoning-uva-2.shinyapps.io/RM_ANOVA",
  # "https://statisticalreasoning-uva-2.shinyapps.io/MultiRegression"
  
)

topics <- gsub("shinyapps.io/", "", gsub("shinyapps.io/", "", str_extract(allLinks, "shinyapps.io/(.*)")))

allLinks <- sprintf('<a href="%s" target="_blank" class="external">link</a>',allLinks) 

knitr::kable(dplyr::arrange(data.frame(Topic = topics, Link = allLinks)), escape = FALSE, format = "html"  ) %>%
  kable_styling(full_width = TRUE)
```

## Additional explanations
```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Same reveal-as-you-go idea as n_lectures_shown in ssr.qmd: uncomment a text
# once the lecture that points students at it has been given. Listed in this
# order, not alphabetically, so they follow the course.
shownTexts <- c(
  "one-sided-vs-two-sided.html",  # lecture 6: correlation
  "eff-size-vs-test-stat.html",   # lecture 6: statistical vs practical significance
  "heteroscedasticity.html"       # lecture 9: linear model assumptions
  # "violating-assumptions-anova.html",
  # "predicting-with-dummies-1-variable.html",
  # "predicting-with-dummies-2-variables.html"
)

# Titles come from each text's own YAML front matter, so the table reads as
# prose rather than as filenames.
textPath  <- "../extra-texts"
htmlFiles <- shownTexts[file.exists(file.path(textPath, sub("\\.html$", ".qmd", shownTexts)))]

titles <- vapply(htmlFiles, function(f) {
  rmarkdown::yaml_front_matter(file.path(textPath, sub("\\.html$", ".qmd", f)))$title
}, character(1))

allLinks <- sprintf('<a href="https://johnnydoorn.github.io/teaching-statistics/extra-texts/%s" target="_blank" class="external">link</a>', htmlFiles)

linkList <- data.frame(Topic = titles, Links = allLinks)

knitr::kable(linkList, escape = FALSE, format = "html", row.names = FALSE) %>%
  kable_styling(full_width = TRUE)

```

## Data + results files
```{r, echo=FALSE, warning=FALSE, message=FALSE}
# One row per data set, driven by the .csv files: the data usually lands before
# the worked results do, so a data set appears as soon as its csv is here and
# the Results/HTML columns fill in later.
dataDir <- paste0("../datasets/", year_dir)
base    <- paste0("https://johnnydoorn.github.io/teaching-statistics/datasets/", year_dir, "/")

present <- list.files(dataDir, full.names = FALSE, recursive = FALSE)
stems   <- sort(sub("\\.csv$", "", grep("\\.csv$", present, value = TRUE)))

link <- function(file) {
  out <- sprintf('<a href="%s%s" target="_blank" class="external">link</a>', base, file)
  out[!(file %in% present)] <- "-"
  out
}

url <- function(href) {
  out <- sprintf('<a href="%s" target="_blank" class="external">link</a>', href)
  out[is.na(href) | href == ""] <- "-"
  out
}

# Columns that live elsewhere, for the book's own data sets: album sales has its
# csv here (the lecture reads it) but the worked results only exist on
# discoverjasp. Named columns override the local link; a topic with no csv here
# gets a row of its own.
externalSets <- list(
  "album_sales" = c(
    Results = "https://discoverjasp.com/repository/dsj1_data/results/album_sales.jasp",
    HTML    = "https://discoverjasp.com/repository/dsj1_data/html/album_sales"
  ),
  # Field's data, hosted by discoverjasp in all three columns, so there is no
  # copy of it here at all. The lecture still reads a csv (discoverjasp serves
  # .jasp only), but that single copy lives in datasets/ and is not duplicated
  # per year.
  "puppies" = c(
    Data    = "https://discoverjasp.com/repository/dsj1_data/data/puppies.jasp",
    Results = "https://discoverjasp.com/repository/dsj1_data/results/puppies.jasp",
    HTML    = "https://discoverjasp.com/repository/dsj1_data/html/puppies.html"
  )
)

dataRows <- data.frame(Topic   = stems,
                       Data    = link(paste0(stems, ".csv")),
                       Results = link(paste0(stems, ".jasp")),
                       HTML    = link(paste0(stems, ".html")))

for (topic in names(externalSets)) {
  set <- externalSets[[topic]]
  if (!topic %in% dataRows$Topic) {
    dataRows <- rbind(dataRows, data.frame(Topic = topic, Data = "-", Results = "-", HTML = "-"))
  }
  dataRows[dataRows$Topic == topic, names(set)] <- as.list(url(set))
}
dataRows <- dataRows[order(dataRows$Topic), ]

if (nrow(dataRows) == 0) {
  knitr::asis_output('<p class="text-muted fst-italic">Data sets will appear here as the semester progresses.</p>')
} else {
  knitr::kable(dataRows, escape = FALSE, format = "html", row.names = FALSE) %>%
    kable_styling(full_width = TRUE)
}
```


## JASP help

The [JASP Video Library](https://jasp-stats.github.io/jasp-video-library/) offers screen-recordings that demonstrate all sorts of cool analyses and features in JASP.

<!-- ```{r child="../jasp-gifs/jasp-gifs-overview.qmd"} -->
<!-- ``` -->

## AI Statement
I use an AI agent to assist with the formatting of the slides and this page. The content is generated by regular I.
