---
title: "Scientific & Statistical Reasoning"
subtitle: "Statistics materials: slides, applets, datasets & extra explanations"
author: "Johnny van Doorn"
date: today
format:
  html:
    theme:
      light: [cosmo, overview-styles.scss]
      dark: [darkly, overview-styles.scss]
    toc: true
    toc-location: left
    toc-title: "On this page"
    title-block-banner: "#C34A5E"
    title-block-banner-color: white
    smooth-scroll: true
    anchor-sections: true
    link-external-newwindow: true
---

::: {.lead-intro}
This page is continuously updated with everything you need for the course: the lecture slides and handouts, interactive applets, the JASP datasets and result files, and extra written explanations. Use the toggle in the sidebar to switch between light and dark mode.
:::

## Slides

```{r child="../ssr/2026/ssr.qmd"}
```


## Applets
```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(stringr)
library("lubridate")
library("dplyr")
library('kableExtra')
allLinks <- c(
   "https://statisticalreasoning-uva.shinyapps.io/NHST_Binomial",
   "https://statisticalreasoning-uva.shinyapps.io/NHST_Continuous"
  # "https://statisticalreasoning-uva.shinyapps.io/NHST_Fdistribution",
  # "https://statisticalreasoning-uva.shinyapps.io/onewayANOVA",
  # "https://statisticalreasoning-uva.shinyapps.io/ANCOVA",
  # "https://statisticalreasoning-uva-2.shinyapps.io/Multi-ANOVA",
  # "https://statisticalreasoning-uva-2.shinyapps.io/RM_ANOVA",
  # "https://statisticalreasoning-uva-2.shinyapps.io/SimpleRegression",
  # "https://statisticalreasoning-uva-2.shinyapps.io/MultiRegression"
  
)

topics <- gsub("shinyapps.io/", "", gsub("shinyapps.io/", "", str_extract(allLinks, "shinyapps.io/(.*)")))

allLinks <- sprintf('<a href="%s" target="_blank" class="external">link</a>',allLinks) 

knitr::kable(dplyr::arrange(data.frame(Topic = topics, Link = allLinks)), escape = FALSE, format = "html"  ) %>%
  kable_styling(full_width = TRUE)
```

## Additional explanations
```{r, echo=FALSE, warning=FALSE, message=FALSE}
files <- list.files(full.names = FALSE, pattern = "\\qmd", recursive = FALSE, path = "../extra-texts")

lectures <- data.frame()
n <- length(files)

files <- files[order(as.numeric(gsub("\\..*", "", files)))]
files <- files[files!= "overview-ssr.qmd"]
htmlFiles <- gsub(x =files, pattern = ".qmd", replacement = ".html")
htmlFiles <- htmlFiles[htmlFiles %in% c(
  # "one-sided-vs-two-sided.html","heteroscedasticity.html", "eff-size-vs-test-stat.html",
  #"violating-assumptions-anova.html", "predicting-with-dummies-1-variable.html",
  #"predicting-with-dummies-2-variables.html"
  )]

allLinks <- sprintf('<a href="https://johnnydoorn.github.io/teaching-statistics/extra-texts/%s" target="_blank" class="external">link</a>',htmlFiles) 

titles <- gsub(htmlFiles, pattern = ".html", replacement = "")

linkList <- data.frame(Topic = titles, Links = allLinks)

knitr::kable(dplyr::arrange(linkList), escape = FALSE, format = "html"  ) %>%
  kable_styling(full_width = TRUE)

```

## Data + results files
```{r, echo=FALSE, warning=FALSE, message=FALSE}
# One row per data set, driven by the .csv files: the data usually lands before
# the worked results do, so a data set appears as soon as its csv is here and
# the Results/HTML columns fill in later.
dataDir <- paste0("../datasets/", year_dir)
base    <- paste0("https://johnnydoorn.github.io/teaching-statistics/datasets/", year_dir, "/")

present <- list.files(dataDir, full.names = FALSE, recursive = FALSE)
stems   <- sort(sub("\\.csv$", "", grep("\\.csv$", present, value = TRUE)))

link <- function(file) {
  out <- sprintf('<a href="%s%s" target="_blank" class="external">link</a>', base, file)
  out[!(file %in% present)] <- "-"
  out
}

if (length(stems) == 0) {
  knitr::asis_output('<p class="text-muted fst-italic">Data sets will appear here as the semester progresses.</p>')
} else {
  knitr::kable(
    data.frame(Topic   = stems,
               Data    = link(paste0(stems, ".csv")),
               Results = link(paste0(stems, ".jasp")),
               HTML    = link(paste0(stems, ".html"))),
    escape = FALSE, format = "html") %>%
    kable_styling(full_width = TRUE)
}
```


## JASP help

The [JASP Video Library](https://jasp-stats.github.io/jasp-video-library/) offers screen-recordings that demonstrate all sorts of cool analyses and features in JASP.

<!-- ```{r child="../jasp-gifs/jasp-gifs-overview.qmd"} -->
<!-- ``` -->

## AI Statement
I use an AI agent to assist with the formatting of the slides and this page. The content is generated by regular I.
