---
title: "11. T-tests"
author: "Johnny van Doorn"
date: "29 September 2026"
view: "https://teams.microsoft.com/meet/383013566330381?p=J5cLpg457MUfuWlfF1"
footer: "Scientific & Statistical Reasoning"
format:
  revealjs:
    output-ext: slide.html
    css: ../../ajohnson.css
  html: default
css: ../../ajohnson.css
---

# Block 2: Life is Mean

- Comparing 2 means (t-test)
- Comparing 2 or more means (ANOVA)
- Between subjects vs. Within subjects
- Adding predictors
    
## {.smaller .unlisted .untitled}

<div class="goals">
In this lecture we aim to:

* Introduce the t-test for comparing means
  * One mean
  * Two means, within subjects
  * Two means, between subjects
* What numbers can we look at? Where do they come from?
    
<strong>Reading:</strong>
Chapter 9 (§9.1–9.14)
</div>

```{r child="t-distribution-shapes.qmd"}
```

```{r child="One-sample_t-test.qmd"}
```

```{r child="t-test_paired.qmd"}
```

```{r child="t-test_independent.qmd"}
```

# Closing

## Next Time

- More means!


## Recommended Exercises

- [Exercise 9.1](https://discoverjasp.com/pages/smart_alex#task-9.1)
- [Exercise 9.3](https://discoverjasp.com/pages/smart_alex#task-9.3)
- [Exercise 9.4](https://discoverjasp.com/pages/smart_alex#task-9.4)
- How would conclusions/interpretations differ for α = 0.01?
- How are the assumptions looking for these data sets?

```{r child="labcoat-leni.qmd"}
```

```{r child="../../footer.qmd"}
```
