---
title: "16. RM ANOVA"
subtitle: "Repeated & Mixed"
author: "Johnny van Doorn"
date: "15 oct 2025"
footer: "Scientific & Statistical Reasoning"
format:
  revealjs:
    output-ext: slide.html
    css: ../../ajohnson.css
  html: default
css: ../../ajohnson.css
---

<div class="goals">
In this lecture we aim to:

- Introduce RM ANOVA
- Introduce Mixed ANOVA
- Show these in JASP
    
<strong>Reading:</strong>
Chapter 14
</div>


```{r child="../../topics/ANOVA_repeated_one-way/ANOVA_repeated_one-way.qmd"}
```

```{r child="../../topics/ANOVA_repeated_factorial/ANOVA_repeated_factorial.qmd"}
```

```{r child="../../topics/ANOVA_mixed/ANOVA_mixed.qmd"}
```

# Closing

## Recap

- In a repeated measures design we can account for baseline differences, to reduce the overall model error
- With within-subjects predictors, we need to check:
  - Sphericity (equal variances of difference scores)

## Recommended Exercises

::: {}
- [Exercise 14.4](https://discoverjasp.com/pages/smart_alex#task-14.4),  [Exercise 14.5](https://discoverjasp.com/pages/smart_alex#task-14.5)
- To understand mechanics of RM ANOVA, [Exercise 14.1](https://discoverjasp.com/pages/smart_alex#task-14.1)
- No sums of squares or effect size calculations required for the exam
:::

<!-- Footer insert below -->

```{r child="../../footer.qmd"}
```
