---
title: "22. Multiple regression"
author: "Sharon Kinkenberg & Johnny van Doorn"
date: "01 nov 2024"
footer: "Scientific & Statistical Reasoning"
format: 
  html: default
  revealjs: 
    output-ext: slide.html
---

<!-- Add the required topics below -->
## Questionnaire variables {.smaller}
- Nationality
- Bachelor specialization
- Ratings
    - Confidence you chose right bachelors
    - Confidence you chose right specialization
    - Confidence in your stats skills
    - Love of stats
    - Mario skills
    - Playfulness
- Number of SSR hours
- Number of friends talked to, in person, in the past week
- Preferred temperature


## WA 10 Data

- [Salary (partial correlation)](https://johnnydoorn.github.io/teaching-statistics/datasets/Salary.jasp)
- [Health (regression)](https://johnnydoorn.github.io/teaching-statistics/datasets/Health.jasp)

```{r child="../../topics/regression_multiple_predictors/regression_multiple_predictors.qmd"}
```

<!-- Footer insert below -->

```{r child="../../footer.qmd"}
```

