---
title: "3. Statistical Reasoning"
subtitle: "NHST and sampling distributions"
author: "Johnny van Doorn"
date: "4 September 2026"
view: "https://amsuni-my.sharepoint.com/:v:/g/personal/j_b_vandoorn_uva_nl/IQA_GS9r9axwT7gDtDgvIoLxAR2fPa9kkI88Tpdg5JfyHMk?e=KV1BIX&nav=eyJyZWZlcnJhbEluZm8iOnsicmVmZXJyYWxBcHAiOiJTdHJlYW1XZWJBcHAiLCJyZWZlcnJhbFZpZXciOiJTaGFyZURpYWxvZy1MaW5rIiwicmVmZXJyYWxBcHBQbGF0Zm9ybSI6IldlYiIsInJlZmVycmFsTW9kZSI6InZpZXcifX0%3D"
footer: "Scientific & Statistical Reasoning"
format:
  revealjs:
    output-ext: slide.html
    css: ../../ajohnson.css
  html: default
css: ../../ajohnson.css
---

<div class="goals">
In block 1 we aim to:

* Discuss NHST concepts
* Introduce JASP
* Correlation
* Linear model (regression)
    
<strong>Reading:</strong>
Chapters 1–8.8
</div>

--- 

<div class="goals">
In this lecture we aim to:

* Introduce the S in SSR
    * The book, the course, and how to study for it
* Repeat some stats concepts from RMS
    * "Logic" behind common hypothesis testing
    * Four scenario's in statistical decision making
* Sampling distributions
    * Discrete: the binomial distribution
    * Continuous: the t-distribution
    
<strong>Reading:</strong>
Chapters 1, 2, 3 (§1.1–1.8, §2.1–2.9, §3.1–3.8)
</div>

```{r child="course-intro.qmd"}
```

```{r child="NHST_3.qmd"}
```

```{r child="t-distribution.qmd"}
```

# Closing

## Next Time

- JASP
  * Descriptive statistics
  * Data visualizations
- Variance, z-scores, standard error and confidence intervals

## Bored?

::: {}
- [Exercise 1.1](https://discoverjasp.com/pages/smart_alex#task-1.1), [Exercise 1.2](https://discoverjasp.com/pages/smart_alex#task-1.2)
- [Exercise 2.3](https://discoverjasp.com/pages/smart_alex#task-2.3), [Exercise 2.6](https://discoverjasp.com/pages/smart_alex#task-2.6), [Exercise 2.7](https://discoverjasp.com/pages/smart_alex#task-2.7)
- [Exercise 3.4](https://discoverjasp.com/pages/smart_alex#task-3.4)
:::

```{r child="../../footer.qmd"}
```
