---
title: "12. ANOVA"
subtitle: "One-way independent ANOVA"
author: "Johnny van Doorn"
date: "TODO 2026"
footer: "Scientific & Statistical Reasoning"
format:
  revealjs:
    output-ext: slide.html
    css: ../../ajohnson.css
  html: default
css: ../../ajohnson.css
---

<div class="goals">
In this lecture we aim to:

- Introduce ANOVA
- Discuss follow-up tests
- Show ANOVA in JASP
    
<strong>Reading:</strong>
Chapter 11 (§11.1–11.13)
</div>


```{r child="ANOVA_one_way_puppies.qmd"}
```


# Closing

## Recap

- ANOVA is a regression (linear model) with categorical predictor
- We compare variance explained by the model, to unexplained variance
- Follow-up tests (contrast/post hoc) inform us where the group differences lie

> bigger group differences = more informative predictor = more explained variance

## Recommended Exercises

- [Exercise 11.1](https://discoverjasp.com/pages/exercises#chapter-11)
- [Exercise 11.2](https://discoverjasp.com/pages/exercises#chapter-11)
- [Exercise 11.4](https://discoverjasp.com/pages/exercises#chapter-11)
- How would conclusions/interpretations differ for α = 0.01?
- How are the assumptions looking for these data sets?

```{r child="../../footer.qmd"}
```
